# (PART) Describing and summarising data  {-}

# Describing variables



```{r echo=FALSE}
SixSteps(4, "Descriptions")
```



## Quantitative and qualitative variables


### Introduction

Before starting on any analysis of the data,
it is important to describe and understand the data,
beacuse the type of data determines how we proceed with summaries and analysis.

Broadly, 
variables may be described as:

* *Quantitative*, 
   which are mathematically numerical;  or 
* *Qualitative*, 
   which are labels or categories. 

Be careful: 
Numerical data does not necessarily mean that the variable is quantitative.
*Mathematically numerical data are quantitative*;
that is, 
numbers that have numerical meanings.
      
```{example}
Postcodes are numbers,
but are *not* quantitative.
The numbers are just labels.
A postcode of 4556 isn't one 'better' or one 'more' than a postcode of 4555.
The values do not have numerical meanings.
Indeed,
rather than numerical postcodes,
alphabetic postcodes could have been chosen.
```








### Quantitative data: Discrete and continuous variables

Quantitative variables may be defined as *discrete* or *continuous*.
Make sure to give the measurement units (such as *kg* or *cm*)
for quantitative data when appropriate.


```{definition, name="Discrete"}
A *discrete* variable is a variable for which the possible values can be counted, at least in theory.
```

Sometimes,
the possible values may not have a theoretical upper limit,
yet can be still considered 'countable'.


```{example}
These (quantitative) variables are discrete:

* The number of heart attacked experience by women in over 40 in the previous year.
The possible values are 0, 1, 2,..
* The number of cracked eggs in a carton of 12. The possible values can be counted: 0, 1, 2, \dots 12.
* Whether or not a person wears an orthic device. The possible values are 'Yes' and '`'No'. 

```


```{definition, name="Continuous variable"}
A *continuous* variable is a variable where all the possible values cannot be counted; 
the number of possible values is infinite,
```

In other words,
there are always values between any two given values of the variable,
if we measure to a greater number of decimal places.
We need to round off to a reasonable value though in practice.


```{example}
These variables are continuous:

* The height of 6-year-old Australian children. 
Between any two given values of heights,
we can find another value by measuring to more and more decimal places of a centimetre,
but we would normally just quote height to the nearest centimetre.
*  The energy consumption of houses in a given city.
Between any two given energy consumption values,
we can find another value by measuring to more and more decimal places of a kiloWatt-hour (kWh),
but we would normally just quote to the nearest kWh.
* The time spent in front of a computer each day for employees in a given industry.
Between any two given times,
we could find another value by measuring to more and more decimal places of a second,
but we would normally just quote the times to the nearest minute, or the nearest 15 minutes.

```

   
   
   
```{example}
Be clear in your definitions!
Consider the variable 'Age'.
	Age is a *continuous* quantitative variable,
	since we age continuously.
	We round our age to the last birthday we  had,
	but that is for convenience.
	(For example, young children's ages may be given as 
	'3 days' or '10 months').

Sometimes we need *Age group* instead
(such as
Under 20; 20 to under 30; 30 to under 50; 50 or over).
Ensure your RQ is clear about which is needed!
```
	

```{exercise}
Here are two survey questions that produce qualitative data.
*How* are the responses different?
*How* are the responses similar?


1. What is your blood type?
    - Type A.
    - Type B
    - Type AB
    - Type O
2. What is your age group?
    - Under 20.
    - 20 to under 30.
    - 30 to under 50.
    - 50 or over.

```      



### Qualitative variables: Nominal and ordinal variables

The possible values of qualitative variables are called 
the *levels* of the variable.
*Nominal variables are qualitative variables where 
the levels have *no natural order*.
*Ordinal* variables are qualitative variables where 
the levels do have a *natural order*.




```{example}  
This survey question will produce *nominal* data:

How do you usually get to university?

- Car (as driver or passenger).
- Bus.
- Ride bicycle or walk.
- Other.

The data will be nominal with four levels.
We could place the levels in any order we want:
for example,
from largest group to smallest,
or in alphabetical order.
We can select the best order for our intended purpose.
```


```{example}
This survey question will produce a *ordinal* data:

Please indicate the extent to which you agree or disagree with this statement:
'Permeable pavements technology has the potential to
revolutionise green building practices'.

- Disagree.
- Neither agree or disagree.
- Agree.

The data will be ordinal with three levels.
It makes sense to treat the levels
in the given order (or the reverse order).
```




```{exercise}
Are these variables
quantitative (discrete or continuous), or
qualitative (nominal or ordinal, with what levels)?
   
* Systolic blood pressure.
* Program of enrolment.
* Academic grade (HD; DN; CR; PS; FL).
* Number of times people visited the doctor last year.

```









## Describing variables in SPSS
   
In SPSS, 
quantitative variables are called *scale*
(regardless of whether they are discrete or continuous).
Nominal variables are called *nominal* in SPSS,
ordinal variables are called *ordinal* in SPSS.
This information is placed in the `Variable View` worksheet of SPSS.



IMAGE








# Graphical summaries of data

```{r echo=FALSE}
SixSteps(4, "Graphical summaries")
```


## Introduction

In any study,
data are collected,
because
**data are the means by which the research question is answered**.
But before analysis,
we need to understand and summarise the data.
In this chapter,
we look at using graphs to summarise data.
Importantly:

>	The purpose of a graph is to display the information
>	in the clearest, simplest possible way,
>	to help us understand the data.

For this reason,
we should always consider what the graph tells us.



## One quantitative variable


Three different types of graphs can be used 
to show how the values of one
quantitative variable are *distributed*.
These are  best used for *continuous* quantitative data,
but may also be used for *discrete* quantitative data
with many possible responses
(such as the number of cars parked in a large shopping centre carpark at hour for year).
The three options are:

* *Stemplot*
(or *stem-and-leaf plot*): 
Best for small amounts of data.
* *Histogram*:
Best for large amounts of data.
* *Dot chart*:
Best for small amounts of data;
good for moderate amounts of data.



### Stem-and-leaf plots

*Stem-and-leaf plots* (or just *stemplots*)
are best described using an example.


```{r echo=FALSE}
BB <- structure(list(Time = c(5L, 104L, 118L, 155L, 257L, 405L, 407L, 
422L, 431L, 708L, 735L, 812L, 814L, 909L, 1035L, 1049L, 1053L, 
1133L, 1209L, 1256L, 1305L, 1406L, 1407L, 1433L, 1446L, 1514L, 
1631L, 1657L, 1742L, 1807L, 1825L, 1854L, 1909L, 1947L, 1949L, 
1951L, 2010L, 2037L, 2051L, 2104L, 2123L, 2217L, 2327L, 2355L
), Gender = c(1L, 1L, 2L, 2L, 2L, 1L, 1L, 2L, 2L, 2L, 2L, 2L, 
1L, 1L, 2L, 1L, 1L, 2L, 2L, 2L, 2L, 1L, 1L, 1L, 1L, 2L, 2L, 2L, 
1L, 2L, 1L, 2L, 2L, 2L, 2L, 2L, 1L, 2L, 2L, 2L, 2L, 1L, 1L, 1L
), Weight = c(3837L, 3334L, 3554L, 3838L, 3625L, 2208L, 1745L, 
2846L, 3166L, 3520L, 3380L, 3294L, 2576L, 3208L, 3521L, 3746L, 
3523L, 2902L, 2635L, 3920L, 3690L, 3430L, 3480L, 3116L, 3428L, 
3783L, 3345L, 3034L, 2184L, 3300L, 2383L, 3428L, 4162L, 3630L, 
3406L, 3402L, 3500L, 3736L, 3370L, 2121L, 3150L, 3866L, 3542L, 
3278L), Minutes = c(5L, 64L, 78L, 115L, 177L, 245L, 247L, 262L, 
271L, 428L, 455L, 492L, 494L, 549L, 635L, 649L, 653L, 693L, 729L, 
776L, 785L, 846L, 847L, 873L, 886L, 914L, 991L, 1017L, 1062L, 
1087L, 1105L, 1134L, 1149L, 1187L, 1189L, 1191L, 1210L, 1237L, 
1251L, 1264L, 1283L, 1337L, 1407L, 1435L)), .Names = c("Time", 
"Gender", "Weight", "Minutes"), class = "data.frame", row.names = c(NA, 
-44L))
```

```{example}
Consider these data,
which are weights (in kg) of babies born 
in a Brisbane hospital in one day
(@mypapers:Dunn:dataset:1999, @data:Steele:BabyBoom).


For these data, 
the weights are given number of kilograms, 
followed by one decimal place of a kilogram.
The whole number of kilograms is placed to the left (as a *stem*),
and the first decimal place is placed on the left (as a *leaf*).
Stem-and-leaf plots work best with the first *two* digits of the data only.
```




```{r echo=FALSE}
htmltools::tags$iframe(title = "Stemplots: Baby weights", src = "./Animations/StemPlotBBMovie.html", height=640, width=520) 
```

\animategraphics[loop,controls, width=\linewidth]{12}{./_book/Animations/images/StemPlotBB}{1}{15}


From this plot,
we see that most births are 3-point-something kilograms.

For stem-and-leaf plots:

* The larger unit (e.g. kilograms) are placed on the left (stems).
* The next smallest unit (e.g. first decimal place of a kilogram) are placed on the right (leaves).
* The numbers in each row should be evenly spaced, so we can see patterns.
* Some data do not work well with stem-and-leaf plots.
  Some times,
  data might need to be suitably rounded before creating the stem-and-leaf plot.
* Within each row, the observations are *ordered* on each stem so patterns can be seen.

   

```{example}
A study of krill [@data:Greenacre2016:reporting] produced the stem-and-leaf plot
shown in the animation below.
This shows that the number of eggs is usually small,
but there are occasional times when a large number of eggs are seen.
```

<!-- ```{r EggsStem, echo=FALSE, fig.cap="Stemplot of the number of eggs in an experiment on *Calanus glacialis*"} -->
<!-- Eggs <- c(1, 0, 1, 2, 2, 8, 16, 31, 0, 0, 20, 0, 26, 1, 3) -->

<!-- # stem(Eggs) -->

<!-- plot.new() -->
<!-- tmp <- capture.output( stem(Eggs)) -->
<!-- text( 0, 1, paste(tmp, collapse='\n'), adj=c(0,1), family='mono') -->
<!-- ``` -->


```{r echo=FALSE}
htmltools::tags$iframe(title = "Stemplots", src = "./Animations/StemPlotMovie.html", height=640, width=520) 
```




### Dot charts


Dot charts show the original data on an axis,
which each observation represented by a point.
   
```{example}
A study examined the serving size of fries at McDonald [@data:Wetzel2005:McDonalds],
and produced the dot chart in Fig \@ref(fig:Fries)
(based on @data:Wetzel2005:McDonalds, Fig. 2).
We can see that the mass of fries is almost always under the target,
and often substantially so.
```

```{r Fries, echo=FALSE, fig.height=3, fig.cap="Mass measurements for large orders of french fries"}
#Fries <- c(58, 60.5, 60.5, 61, 62, 62, 62.5, 64, 67, 67, 69, 69, 71, 71, 71, 71.5, 
#           76, 76.5, 76.5, 77, 77, 77.5, 79.5, 82, 82, 82.5, 84, 84, 87, 93, 104)
#stripchart(Fries, 
#           method="stack",
#           main="Dotplot for Mass of Small Order of Fries",
#           axes=FALSE,
#           ylim = c(-5, 10),
#           xlim=c(65, 110) )
#axis(side=1)
#arrows(74, 8, 74, 2, angle=15)
#text(74, 8, "McDonald's target of 74 grams per serving", cex=0.8,pos=3)


Fries <- c(117, 126, 128, 132, 133, 133, 134, 137, 138, 139, 139, 140, 141, 142, 142.5, 143, 143.5,  
           145, 146, 146, 151, 152, 152, 154, 154, 154.5, 154.5, 155, 156, 156.5, 157, 176)
stripchart(Fries, 
           method="stack",
           main="Dotplot for Mass of Large Order of Fries",
           axes=FALSE,
           pch=20,
           ylim = c(-1, 10),
           xlim=c(112, 180) )
axis(side=1, 
     at= seq(115, 175, by=10) )
arrows(165, 8, 171, 2, angle=15)
text(165, 8, 
     "McDonald's target of 171 grams per serving", 
     cex=0.8,
     pos=3)
```


### Histograms

Histograms are a series of boxes,
where the width of the box represents a range of values of the variable being graphed,
and the height of the box represents the number (or percentage) of observations within that range of values.
The boxes on a histogram show how many observations
lie within given ranges of the quantitative variable:

```{example}
Consider again the
weights (in kg) of babies born 
in a Brisbane hospital in one day
(@mypapers:Dunn:dataset:1999, @data:Steele:BabyBoom).
A histogram can be constructed, 
as shown in the animation below.
```


```{r echo=FALSE}
htmltools::tags$iframe(title = "Histogram: Baby weights", src = "./Animations/BabyboomHistMovie.html", height=640, width=520) 
```


```{example}
A study of 'headache attributed to ingestion or inhalation of a cold stimulus' (HICS),
commonly known as a brain freeze from eating cold food (e.g. ice cream) or drinking cold drink,
measured the duration of the brain freeze [@data:Mages2017:BrainFreeze].

A histogram of the data (Fig \@ref(fig:BrainFreeze),
based on @data:Mages2017:BrainFreeze, Figure 2b).
shows that
11 people experience HICS symptoms between 0 and 5 second in length.
In addition, 9 people experienced symptoms between 5 and 10 seconds,
and 1 person experienced symptoms for more than 35 seconds.
```

```{r BrainFreeze, echo=FALSE, fig.cap="Duration of HICS after drinking ice water"}
# Read from Mages 2017, Figure 2b

Hts <- c(11, 9, 4, 4, 2, 5, 0, 1)
Borders <- seq(0, 40, by=5)

old.par <- par()


barplot(Hts, 
        space=0,
        ylim=c(0, 12),
        col=plot.colour,
        main="Duration of HICS after drinking ice water",
        ylab="Number of volunteers",
        xlab="Duration (s)",
        axes=FALSE)
axis(side=2, 
     las=1)
axis(side=1, 
     at=0:8, 
     labels=Borders)
box()
```


   




### Summarising the data {#SummariseData}

Recall:
Graphs are constructed to help us understand the data.
After producing a graph for one quantitative variable, then,
we need to summarise what we learn.
For one *quantitative variable*,
describe:

1. Location:
   What is an "average", central or typical value?
2. Variation: 
    How much variation is present in the bulk of the data?
3.  Shape:
    How are the values distributed?
4.  Outliers and unusual features:
    Are there any unusual observations, or anything else of interest?

Describing the shape can be tricky,
but some terminology helps
(Fig. \@ref(fig:DescribeShape)):

*	Skewed *right*;
*	Skewed *left*;
* Symmetric data (and perhaps bell-shaped)
* Bimodal data.

Sometime, there are no suitable short descriptions.

```{r DescribeShape, echo=FALSE, fig.height=8, fig.cap="Some common shapes of the distribution of qualitative data"}
par(mfrow=c(2,2))

library(scales)

set.seed(110110)

xx <- c( rchisq(300, df=2), 7,7,8,7,7,7,8,7,7,6)

hist( scales::rescale(xx, to=c(0, 5)), 
   xlab="Values", ylab="Number of obs.",
   main="Skewed right",
   col="slateblue", las=1)

xx <- c( rchisq(300, df=1), 2, 2, 1, 1, 2, 0, 0)

xx <- max(xx) - xx
hist( scales::rescale(xx, to=c(0, 5)), 
   xlab="Values", ylab="Number of obs.",
   main="Skewed left",
   col="slateblue", las=1)

xx <- rnorm(500, mean=10)
hist( scales::rescale(xx, to=c(0, 10)), 
   xlab="Values", ylab="Number of obs.",
   main="Approx. symmetric",
   col="slateblue", las=1)


xx1 <- rnorm(500,mean=0, sd=1)
xx2 <- rnorm(250,mean=4, sd=1)
xx <- c( xx1[xx1>0], xx2)


hist( scales::rescale(xx, to=c(0, 10)),  
	right=TRUE,
   col="slateblue", las=1,
   xlab="Values",ylab="Number of obs.", main="Bimodal distribution")

par(mfrow=c(1,1))
```



```{exercise}
Describe the histogram in
Example \@ref(fig:DescribeShape).
```













## One qualitative variable


Three different graphs can be used to show how common 
each level of a qualitative variable is in the data.
These are best used for qualitative data,
are quantitative datas with a small number of possible options 
may also be graphed using these graphs
(e.g. number of cars for sale on an online selling site with 4, 6 and 8 cylinders).
The three options are:

* *Dot chart*:
   Usually a good choice.
* *Bar chart*:
   Usually a good choice.
* *Pie chart*:
   Only useful in special circumstances;
   can be harder to interpret.

For *nominal* data, 
order is unimportant;
so we could order the categories 
alphabetically, by size;
by my preference;
or any way I like.

For ordinal data,
the natural order of the levels should be used.



### Dot charts

Dot charts indicate the counts (or the corresponding percentages)
in each level
using dots.
The levels can be on the horizontal or vertical axis;
placing the level names on the vertical axis often makes for easier reading,
and more room for long labels. 



```{example}
A study of spider monkeys [@data:Chapman1990:SpiderMonkeys]
examined the types of social groups.
A dot chart (Fig. \@ref(fig:SpiderMonkeys)) can be used to display the data.
```


```{r SpiderMonkeys, echo=FALSE, fig.cap="2D dotchart of spider monkey family groups"}
counts <- c(8, 3, 2, 15, 1, 23, 48)
names <- c("Solitary", "All males", "Female + no young", "Mixed young", 
           "Mixed + no young", "One female + offspring", "Many females + offspring")

sort.order <- sort( counts, index.return=TRUE)

old.mar=par()$mar
new.mar <- old.mar
new.mar[2] <- old.mar[2]*2.5
par(mar = new.mar  )

dotchart( counts[sort.order$ix], 
          pch=19,
          las=1, 
          xlim=c(0,50),
          xlab="Percentage of observations",
          ylab="",
          main="Spider monkey sub-groups")
axis(side=2, seq_along(counts[sort.order$ix]), names[sort.order$ix], las=1)

```




### Bar charts

Similar to dot charts,
bar charts indicate the counts in each category
using a bar from the zero-line.
As with dot charts,
the levels can be on the horizontal or vertical axis,
and placing the level names on the vertical axis often makes for easier reading,
and more room for long labels. 


```{example IndependenceBarText}
In a study of functional independence [@data:Ocepek2013:SmartHome],
the type of diagnoses were graphed using a bar chart:
(Fig. \@ref(fig:IndependenceBar))
``` 

```{r IndependenceBar, echo=FALSE, fig.cap="Diagnoses of participants"}
Number <- rev( c(11, 10, 9, 8, 5, 4, 3, 3, 2, 4) )
Diag <- rev( c("Amputation-one leg",
          "Neuromuscular disease",
          "SCI-paraplegia",
          "Rheum. dis. or multiple skeletal injury",
          "SCI-tetraplegia",
          "Cerebral vascular insult",
          "Amputation-both legs",
          "Amputation-arm",
          "Cerebral palsy",
          "Other") )
Diag.Cols <- rev( c( 
                grey(0.6), 
                grey(0.2), 
                grey(0.6),
                grey(0.6),
                grey(0.2),
                grey(0.6),
                grey(0.2),
                rep( grey(0.6), 3) ) )

par( mar=c(4, 16, 4, 2) + 0.1)
barplot( Number, 
         names.arg=Diag,
         horiz=TRUE,
         las=1,
         xlim=c(0, 12),
         xlab="Number of participants",
         col=Diag.Cols)
```


For bar charts and for dot charts:

* Place the qualitative variable on the horizontal or vertical axis 
   (and label with the levels of the variable).
* Use a count or percentage on the other axis.
*  For nominal data,
   dots and bars can be ordered any way we like:
   *Think about the most helpful order*.
* Bars have gaps between bars, as the bars represent distinct categories.
In contrast, the bars of histograms are butted together
(exept when an interval has a coun of zero),
as the bars represent a numerical scale.









### Pie charts

In pie charts,
a circle is divided into segments 
proportional to the number of counts.

```{example}
In a study of functional independence [@data:Ocepek2013:SmartHome],
the severity of the  diagnoses were graphed using a pie chart
(\@ref(fig:IndependencePie)).

This picture actually conveys one thing only:
"69\% of patients had a less severe injury".
Notice that the colour coding helps explain the colours
used in Example \@ref(exm:IndependenceBarText).
```
   
   
```{r IndependencePie, echo=FALSE, fig.cap="Severity of diagnoses of participants"}
pie( c(31, 69),
     labels=c("More severe (31%)", "Less severe (69%"),
     col=c( grey(0.2), grey(0.6)),
     init.angle = 90)
```


Pie charts need *all* categories available,
and display *parts of a whole* (so they must add to 100\%),
so not all qualitative data can be graphed with a pie chart.
In addition,
categories with zero observations cannot be shown in a pie chart.
Pie charts are also difficult to produce and understand for variables
with many categories.



```{exercise}
In which of these situations is a pie chart *appropriate*?
  
* The percentage of users of these web browsers: Firefox, Chrome, and Safari.
* The percentage of people in each state who own an iPhone.
* The percentage of students awarded different grades last semester.

```



```{example}
A study of patients with thrombolysis [@data:Castle206:Thrombolysis]
produced the following pie chart:
	

This chart conveys one number:
"14.29\% of the patients were suitable for the study".
```









### Comparing pie charts and bar charts

Consider the pie chart
(made from data in @data:houchens:1988)
in the left panel of
Fig. \@ref(fig:LOSgraphs).
A pie chart *is* appropriate,
as the counts are parts of a whole.
But notice that comparing the sizes of the segments is hard.
Compare this to the equivalent bar chart in the right panel.
In general,
human brains are better at comparing *lengths*
than they are at comparing *angles* [@data:Friel:Graphs].






```{r LOSgraphs, echo=FALSE, fig.cap="Graphs from a study of hispotal admission of children with asthma"}
los <- 
structure(list(LOS = c(1L, 2L, 3L, 4L, 2L, 2L, 1L, 1L, 2L, 2L, 
3L, 1L, 1L, 1L, 2L, 1L, 1L, 3L, 2L, 3L, 1L, 2L, 1L, 2L, 4L, 2L, 
3L, 5L, 2L, 6L, 3L, 1L, 1L, 3L, 1L, 2L, 4L, 1L, 3L, 7L, 2L, 4L, 
4L, 1L, 3L, 3L, 3L, 2L, 1L, 2L, 3L, 2L, 6L, 3L, 2L, 3L, 1L, 2L, 
1L, 3L, 3L, 3L, 1L, 1L, 2L, 2L, 2L, 3L, 2L, 2L, 3L, 3L, 2L, 3L, 
1L, 2L, 2L, 3L, 1L, 4L, 1L, 5L, 2L, 1L, 3L, 1L, 1L, 2L, 2L, 2L, 
1L, 2L, 2L, 2L, 1L, 3L, 1L, 1L, 2L, 2L, 2L, 1L, 3L, 1L, 1L, 1L, 
3L, 2L, 1L, 2L, 3L, 2L, 2L, 2L, 2L, 2L, 1L, 2L, 3L, 3L, 1L, 4L, 
1L, 4L, 3L, 4L, 3L, 2L, 2L, 1L, 2L, 8L, 2L, 2L, 2L, 4L, 4L, 2L, 
2L, 1L, 2L, 3L, 3L, 2L, 2L, 2L, 1L, 1L, 2L, 3L, 4L, 3L, 4L, 4L, 
2L, 2L, 3L, 2L, 1L, 2L, 1L, 2L, 2L, 2L, 1L, 1L, 2L, 2L, 3L, 2L, 
2L, 2L, 3L, 3L, 3L, 2L, 2L, 2L, 3L, 2L, 2L, 2L, 2L, 3L, 1L, 4L, 
3L, 1L, 2L, 2L, 2L, 2L, 2L, 1L, 4L, 2L, 3L, 2L, 2L, 1L, 3L, 2L, 
3L, 3L, 1L, 2L, 2L, 3L, 1L, 1L, 2L, 1L, 1L, 1L, 2L, 1L, 5L, 2L, 
2L, 3L, 2L, 2L, 3L, 4L, 1L, 2L, 2L, 2L, 1L, 6L, 3L, 3L, 3L, 2L, 
1L, 5L, 1L, 2L, 1L, 2L, 2L, 3L, 1L, 7L, 4L, 2L, 2L, 2L, 3L, 2L, 
2L, 2L, 3L, 1L, 2L, 1L, 3L, 4L, 4L, 3L, 2L, 1L, 3L, 3L, 3L, 3L, 
3L, 4L, 2L, 2L, 3L, 2L, 2L, 1L, 3L, 1L, 1L, 2L, 2L, 3L, 3L, 2L, 
2L, 3L, 6L, 3L, 2L, 5L, 3L, 4L, 4L, 3L, 4L, 2L, 4L, 2L, 2L, 2L, 
2L, 4L, 2L, 5L, 1L, 2L, 3L, 3L, 1L, 2L, 4L, 2L, 1L, 1L, 3L, 3L, 
2L, 3L, 10L, 2L, 3L, 2L, 2L, 4L, 2L, 4L, 1L, 3L, 4L, 3L, 3L, 
4L, 3L, 1L, 3L, 3L, 5L, 2L, 2L, 3L, 1L, 2L, 3L, 1L, 1L, 1L, 1L, 
2L, 2L, 3L, 1L, 1L, 2L, 2L, 2L, 1L, 2L, 1L, 2L, 1L, 1L, 1L, 3L, 
1L, 2L, 1L, 2L, 6L, 2L, 2L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 
2L, 3L, 3L, 2L, 2L, 5L, 4L, 5L, 2L, 1L, 2L, 1L, 2L, 3L, 3L, 2L, 
2L, 3L, 2L, 3L, 2L, 1L, 4L, 1L, 4L, 1L, 1L, 1L, 6L, 2L, 2L, 1L, 
5L, 2L, 4L, 3L, 1L, 3L, 1L, 3L, 3L, 7L, 1L, 1L, 1L, 3L, 1L, 5L, 
2L, 5L, 2L, 3L, 4L, 2L, 2L, 1L, 2L, 5L, 2L, 3L, 2L, 3L, 2L, 1L, 
1L, 1L, 3L, 1L, 3L, 1L, 2L, 2L, 3L, 2L, 5L, 4L, 2L, 5L, 2L, 2L, 
4L, 3L, 2L, 1L, 1L, 6L, 1L, 9L, 2L, 1L, 6L, 3L, 3L, 2L, 3L, 2L, 
3L, 3L, 3L, 3L, 3L, 3L, 3L, 2L, 3L, 2L, 1L, 4L, 2L, 2L, 2L, 1L, 
2L, 3L, 3L, 2L, 1L, 2L, 2L, 4L, 1L, 2L, 4L, 3L, 1L, 2L, 2L, 1L, 
2L, 4L, 3L, 2L, 2L, 3L, 1L, 2L, 5L, 3L, 2L, 2L, 1L, 4L, 6L, 1L, 
4L, 3L, 2L, 3L, 3L, 3L, 3L, 2L, 5L, 5L, 3L, 2L, 2L, 4L, 2L, 1L, 
2L, 2L, 2L, 3L, 3L, 1L, 3L, 5L, 2L, 2L, 3L, 3L, 2L, 3L, 1L, 1L, 
3L, 2L, 2L, 4L, 2L, 2L, 6L, 4L, 2L, 1L, 3L, 1L, 2L, 3L, 2L, 1L, 
3L, 1L, 2L, 2L, 2L, 2L, 5L, 6L, 4L, 2L, 2L, 5L, 1L, 1L, 2L, 1L, 
3L, 1L, 1L, 1L, 8L, 3L, 3L, 2L, 2L, 2L, 3L, 3L, 7L, 1L, 5L, 6L, 
3L, 4L, 2L, 4L, 4L, 5L, 5L, 4L, 5L, 7L, 3L, 3L, 3L, 3L, 4L, 3L, 
1L, 1L, 3L, 5L, 4L, 4L, 6L, 6L, 6L, 3L, 4L, 1L, 2L, 1L, 1L, 1L, 
2L, 6L, 5L, 4L, 4L, 2L, 4L, 3L, 2L, 2L, 3L, 1L, 3L, 4L, 2L, 4L, 
7L, 2L, 3L, 4L, 2L, 2L, 3L, 2L, 6L, 5L, 7L, 4L, 1L, 2L, 3L, 3L, 
2L, 1L, 5L, 3L, 2L, 2L, 4L, 1L, 5L, 3L, 3L, 5L, 5L, 8L, 2L, 5L, 
2L, 3L, 3L, 2L, 4L, 2L, 2L, 3L, 3L, 6L, 2L, 3L, 3L, 4L, 4L, 4L, 
4L, 2L, 4L, 1L, 2L, 1L, 5L, 4L, 6L, 2L, 5L, 4L, 1L, 8L, 3L, 2L, 
2L, 3L, 2L, 3L, 1L, 2L, 2L, 2L, 1L, 2L, 4L, 2L, 2L, 2L, 2L, 1L, 
2L, 3L, 1L, 6L, 2L, 6L, 2L, 11L, 3L, 4L, 2L, 3L, 3L, 4L, 2L, 
4L, 2L, 4L, 4L, 4L, 3L, 5L, 3L, 2L, 1L, 3L, 1L, 4L, 2L, 2L, 4L, 
3L, 3L, 3L, 3L, 2L, 3L, 4L, 2L, 5L, 4L, 6L, 5L, 3L, 7L, 2L, 3L, 
3L, 4L, 2L, 2L, 4L, 3L, 3L, 6L, 1L, 3L, 2L, 2L, 2L), Hospital = c(17L, 
17L, 17L, 17L, 17L, 17L, 17L, 17L, 17L, 17L, 17L, 17L, 17L, 17L, 
17L, 17L, 17L, 27L, 27L, 27L, 57L, 57L, 57L, 57L, 3L, 3L, 3L, 
3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 
3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 18L, 18L, 18L, 18L, 18L, 18L, 
23L, 23L, 23L, 23L, 23L, 23L, 23L, 23L, 23L, 23L, 23L, 23L, 23L, 
23L, 23L, 23L, 23L, 23L, 23L, 23L, 23L, 23L, 23L, 23L, 23L, 23L, 
23L, 23L, 23L, 7L, 7L, 7L, 30L, 30L, 30L, 30L, 30L, 30L, 30L, 
30L, 30L, 30L, 30L, 24L, 24L, 24L, 24L, 24L, 24L, 24L, 24L, 24L, 
24L, 24L, 24L, 24L, 24L, 24L, 24L, 24L, 24L, 24L, 24L, 24L, 24L, 
24L, 24L, 24L, 24L, 24L, 24L, 24L, 24L, 24L, 24L, 22L, 22L, 22L, 
22L, 22L, 22L, 22L, 22L, 22L, 22L, 22L, 22L, 22L, 22L, 22L, 22L, 
22L, 22L, 22L, 22L, 22L, 22L, 22L, 22L, 22L, 22L, 22L, 22L, 22L, 
22L, 22L, 10L, 10L, 10L, 10L, 10L, 10L, 10L, 10L, 10L, 10L, 10L, 
6L, 6L, 6L, 6L, 6L, 6L, 6L, 6L, 6L, 6L, 6L, 6L, 20L, 20L, 20L, 
20L, 20L, 20L, 20L, 20L, 20L, 20L, 20L, 20L, 20L, 20L, 20L, 20L, 
20L, 20L, 20L, 20L, 20L, 20L, 20L, 20L, 20L, 20L, 20L, 20L, 20L, 
20L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 
49L, 49L, 49L, 49L, 49L, 49L, 49L, 49L, 49L, 49L, 44L, 48L, 48L, 
48L, 48L, 48L, 48L, 48L, 48L, 48L, 48L, 48L, 48L, 48L, 56L, 56L, 
56L, 56L, 56L, 56L, 56L, 56L, 56L, 56L, 56L, 56L, 56L, 1L, 1L, 
1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 9L, 9L, 9L, 9L, 9L, 9L, 9L, 9L, 
9L, 9L, 9L, 2L, 2L, 50L, 50L, 50L, 50L, 50L, 50L, 58L, 58L, 58L, 
58L, 58L, 58L, 58L, 58L, 58L, 58L, 58L, 58L, 58L, 58L, 58L, 58L, 
58L, 58L, 58L, 58L, 58L, 58L, 58L, 58L, 58L, 58L, 45L, 45L, 45L, 
45L, 45L, 45L, 5L, 5L, 5L, 5L, 19L, 19L, 19L, 19L, 19L, 19L, 
19L, 19L, 19L, 19L, 19L, 19L, 19L, 19L, 19L, 19L, 19L, 19L, 19L, 
19L, 19L, 19L, 19L, 19L, 19L, 19L, 19L, 19L, 19L, 19L, 18L, 18L, 
18L, 18L, 18L, 50L, 50L, 50L, 19L, 19L, 19L, 8L, 8L, 8L, 8L, 
17L, 17L, 17L, 17L, 17L, 17L, 17L, 17L, 17L, 17L, 17L, 17L, 17L, 
17L, 27L, 27L, 27L, 27L, 27L, 27L, 27L, 27L, 27L, 27L, 27L, 27L, 
27L, 27L, 27L, 27L, 27L, 27L, 27L, 27L, 27L, 27L, 27L, 27L, 27L, 
27L, 27L, 27L, 27L, 57L, 57L, 57L, 57L, 57L, 57L, 57L, 57L, 57L, 
57L, 57L, 57L, 57L, 57L, 57L, 57L, 57L, 57L, 57L, 2L, 2L, 2L, 
2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 18L, 18L, 18L, 18L, 
18L, 18L, 18L, 18L, 18L, 18L, 18L, 18L, 18L, 18L, 23L, 7L, 7L, 
7L, 7L, 7L, 7L, 7L, 7L, 7L, 7L, 7L, 7L, 7L, 7L, 7L, 7L, 7L, 7L, 
7L, 7L, 7L, 7L, 7L, 7L, 7L, 7L, 7L, 7L, 7L, 7L, 7L, 7L, 7L, 7L, 
30L, 30L, 30L, 30L, 30L, 30L, 30L, 30L, 30L, 30L, 30L, 30L, 30L, 
30L, 30L, 30L, 30L, 24L, 24L, 24L, 10L, 10L, 10L, 4L, 4L, 4L, 
4L, 6L, 6L, 6L, 6L, 6L, 6L, 6L, 6L, 6L, 6L, 20L, 20L, 20L, 20L, 
4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 
4L, 4L, 4L, 4L, 4L, 4L, 4L, 5L, 5L, 5L, 5L, 5L, 5L, 5L, 49L, 
49L, 49L, 49L, 49L, 49L, 49L, 49L, 49L, 44L, 44L, 44L, 44L, 44L, 
44L, 44L, 44L, 44L, 44L, 44L, 44L, 44L, 44L, 44L, 48L, 48L, 48L, 
48L, 48L, 48L, 48L, 48L, 48L, 48L, 48L, 48L, 48L, 48L, 48L, 56L, 
56L, 56L, 56L, 56L, 56L, 56L, 56L, 56L, 56L, 56L, 56L, 56L, 56L, 
56L, 56L, 56L, 56L, 56L, 56L, 56L, 56L, 56L, 56L, 42L, 42L, 42L, 
42L, 42L, 42L, 42L, 42L, 42L, 42L, 42L, 42L, 42L, 42L, 42L, 42L, 
42L, 42L, 42L, 42L, 42L, 42L, 42L, 42L, 1L, 1L, 1L, 1L, 1L, 1L, 
1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 
9L, 9L, 9L, 9L, 9L, 9L, 9L, 9L, 9L, 9L, 9L, 54L, 54L, 54L, 54L, 
54L, 54L, 54L, 54L, 54L, 54L, 54L, 54L, 54L, 54L, 54L, 54L, 54L, 
54L, 54L, 54L, 54L, 50L, 50L, 50L, 50L, 50L, 50L, 50L, 50L, 50L, 
50L, 50L, 50L, 50L, 50L, 50L, 50L, 50L, 50L, 50L, 50L, 50L, 50L, 
50L, 58L, 58L, 45L, 45L, 45L, 45L, 45L, 58L, 58L, 58L, 18L, 18L, 
18L, 18L, 18L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 6L, 6L, 6L, 
6L, 6L, 6L, 6L, 6L, 6L, 6L, 6L, 6L), Insurer = c(0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 
1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 
1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 
1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 
1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 
1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 
1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 
1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 
1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 
1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 
1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 
1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 
1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 
1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 
1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 
1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 
1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 
1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 
1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 
1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 
1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 
1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 0L, 1L, 1L, 
1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 
1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 
1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 
1L, 1L), Age = c(10L, 6L, 3L, 11L, 3L, 13L, 2L, 16L, 12L, 13L, 
4L, 3L, 3L, 7L, 2L, 3L, 4L, 2L, 7L, 15L, 3L, 4L, 2L, 9L, 2L, 
3L, 11L, 2L, 3L, 7L, 6L, 5L, 5L, 8L, 2L, 2L, 2L, 5L, 13L, 9L, 
2L, 15L, 4L, 14L, 17L, 4L, 6L, 7L, 13L, 2L, 3L, 13L, 4L, 16L, 
11L, 13L, 2L, 13L, 6L, 7L, 3L, 5L, 4L, 8L, 4L, 10L, 2L, 5L, 3L, 
6L, 4L, 4L, 4L, 4L, 8L, 4L, 4L, 5L, 10L, 4L, 4L, 8L, 5L, 3L, 
7L, 4L, 8L, 9L, 4L, 2L, 16L, 2L, 2L, 4L, 2L, 3L, 2L, 2L, 9L, 
13L, 13L, 3L, 12L, 9L, 9L, 2L, 5L, 3L, 5L, 6L, 7L, 2L, 4L, 5L, 
3L, 3L, 14L, 13L, 17L, 13L, 10L, 10L, 9L, 15L, 4L, 11L, 8L, 5L, 
4L, 8L, 4L, 2L, 3L, 2L, 14L, 9L, 10L, 3L, 2L, 3L, 11L, 13L, 2L, 
7L, 10L, 11L, 17L, 5L, 4L, 10L, 17L, 15L, 14L, 13L, 12L, 13L, 
13L, 12L, 12L, 11L, 12L, 9L, 11L, 9L, 15L, 11L, 7L, 16L, 8L, 
3L, 9L, 4L, 3L, 3L, 2L, 4L, 3L, 3L, 2L, 4L, 6L, 3L, 2L, 4L, 9L, 
6L, 2L, 13L, 15L, 6L, 3L, 11L, 2L, 2L, 8L, 2L, 2L, 6L, 3L, 3L, 
5L, 3L, 4L, 17L, 5L, 2L, 3L, 5L, 2L, 14L, 15L, 16L, 4L, 3L, 2L, 
13L, 3L, 9L, 2L, 3L, 7L, 11L, 15L, 10L, 8L, 7L, 6L, 13L, 2L, 
15L, 14L, 15L, 10L, 11L, 7L, 9L, 3L, 10L, 2L, 3L, 3L, 15L, 4L, 
16L, 5L, 5L, 6L, 8L, 2L, 7L, 6L, 3L, 5L, 5L, 3L, 2L, 4L, 11L, 
8L, 8L, 16L, 2L, 3L, 9L, 5L, 11L, 5L, 2L, 2L, 5L, 8L, 3L, 4L, 
14L, 8L, 2L, 2L, 17L, 2L, 12L, 8L, 6L, 2L, 17L, 3L, 2L, 5L, 11L, 
3L, 3L, 13L, 3L, 8L, 13L, 13L, 13L, 13L, 3L, 3L, 2L, 4L, 7L, 
5L, 2L, 3L, 6L, 5L, 3L, 2L, 9L, 13L, 3L, 5L, 2L, 11L, 11L, 9L, 
4L, 5L, 5L, 4L, 15L, 3L, 5L, 9L, 13L, 3L, 5L, 6L, 11L, 3L, 2L, 
7L, 2L, 8L, 2L, 4L, 14L, 4L, 12L, 2L, 2L, 2L, 10L, 7L, 3L, 3L, 
9L, 17L, 17L, 16L, 16L, 12L, 11L, 11L, 9L, 14L, 6L, 5L, 5L, 5L, 
5L, 4L, 3L, 4L, 2L, 5L, 2L, 3L, 4L, 11L, 6L, 4L, 3L, 3L, 4L, 
3L, 4L, 11L, 14L, 4L, 8L, 8L, 7L, 7L, 14L, 6L, 3L, 11L, 4L, 8L, 
5L, 2L, 4L, 17L, 4L, 2L, 6L, 3L, 5L, 6L, 11L, 5L, 2L, 10L, 2L, 
3L, 4L, 3L, 13L, 13L, 4L, 5L, 6L, 14L, 6L, 13L, 3L, 4L, 11L, 
13L, 2L, 15L, 3L, 2L, 2L, 2L, 7L, 2L, 4L, 2L, 2L, 5L, 14L, 7L, 
3L, 5L, 12L, 4L, 3L, 2L, 4L, 2L, 15L, 7L, 11L, 15L, 15L, 10L, 
5L, 16L, 6L, 2L, 3L, 4L, 3L, 14L, 5L, 6L, 6L, 5L, 6L, 3L, 14L, 
3L, 3L, 10L, 6L, 11L, 3L, 4L, 2L, 3L, 17L, 17L, 4L, 4L, 4L, 13L, 
2L, 5L, 3L, 3L, 3L, 14L, 7L, 5L, 5L, 13L, 3L, 6L, 11L, 3L, 7L, 
6L, 8L, 4L, 3L, 2L, 12L, 2L, 5L, 6L, 7L, 6L, 6L, 13L, 5L, 6L, 
2L, 6L, 5L, 7L, 14L, 8L, 5L, 8L, 9L, 5L, 2L, 10L, 4L, 4L, 7L, 
2L, 3L, 2L, 2L, 6L, 10L, 10L, 14L, 14L, 2L, 7L, 10L, 3L, 3L, 
2L, 14L, 14L, 10L, 5L, 4L, 2L, 7L, 4L, 4L, 5L, 11L, 5L, 12L, 
3L, 9L, 3L, 3L, 2L, 3L, 8L, 13L, 10L, 5L, 5L, 14L, 5L, 12L, 12L, 
8L, 6L, 6L, 6L, 4L, 5L, 5L, 6L, 3L, 5L, 2L, 4L, 11L, 11L, 2L, 
9L, 3L, 2L, 10L, 3L, 6L, 12L, 17L, 10L, 4L, 4L, 17L, 11L, 2L, 
6L, 15L, 4L, 10L, 5L, 2L, 16L, 6L, 5L, 6L, 4L, 12L, 5L, 4L, 2L, 
5L, 5L, 7L, 2L, 14L, 4L, 14L, 5L, 16L, 4L, 3L, 3L, 17L, 13L, 
6L, 2L, 3L, 3L, 5L, 2L, 7L, 2L, 2L, 13L, 2L, 3L, 3L, 3L, 12L, 
6L, 7L, 8L, 4L, 4L, 13L, 11L, 2L, 2L, 6L, 15L, 14L, 2L, 2L, 3L, 
2L, 9L, 17L, 12L, 11L, 11L, 4L, 3L, 4L, 2L, 2L, 2L, 6L, 2L, 4L, 
2L, 7L, 4L, 5L, 5L, 3L, 4L, 7L, 7L, 7L, 8L, 5L, 8L, 3L, 3L, 2L, 
3L, 2L, 10L, 2L, 16L, 14L, 10L, 6L, 3L, 3L, 8L, 2L, 7L, 12L, 
5L, 3L, 2L, 14L, 15L, 5L, 11L, 13L, 13L, 17L, 12L, 9L, 5L, 5L, 
11L, 4L, 3L, 2L, 11L, 2L, 15L, 2L, 10L, 9L, 3L, 3L, 5L, 6L, 17L, 
4L, 14L, 9L, 14L, 8L, 6L, 5L, 4L, 4L, 4L, 2L, 2L, 2L, 5L, 8L, 
5L, 3L, 3L, 6L, 6L, 3L, 6L, 2L, 5L, 10L, 4L, 5L, 9L, 3L, 9L, 
6L, 4L, 8L, 9L, 4L, 3L, 9L, 3L, 4L, 7L, 4L, 4L, 7L, 7L, 11L, 
4L, 4L, 17L, 10L, 6L, 8L, 10L, 12L, 2L, 6L, 3L, 14L, 2L, 12L, 
12L), Gender = c(1L, 0L, 0L, 0L, 0L, 0L, 1L, 0L, 1L, 0L, 0L, 
1L, 0L, 1L, 0L, 1L, 0L, 1L, 1L, 1L, 0L, 1L, 0L, 1L, 0L, 0L, 0L, 
1L, 0L, 0L, 0L, 1L, 0L, 1L, 0L, 1L, 1L, 1L, 0L, 0L, 0L, 0L, 0L, 
1L, 0L, 0L, 1L, 0L, 1L, 1L, 0L, 1L, 0L, 0L, 0L, 0L, 1L, 0L, 0L, 
0L, 0L, 1L, 0L, 1L, 1L, 1L, 1L, 1L, 0L, 0L, 0L, 0L, 0L, 1L, 1L, 
0L, 0L, 0L, 1L, 1L, 0L, 1L, 1L, 0L, 1L, 0L, 1L, 0L, 0L, 0L, 1L, 
0L, 1L, 1L, 0L, 0L, 0L, 1L, 1L, 0L, 0L, 1L, 0L, 0L, 0L, 1L, 0L, 
1L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 1L, 0L, 0L, 0L, 0L, 1L, 0L, 
0L, 0L, 1L, 0L, 0L, 0L, 1L, 0L, 0L, 0L, 0L, 1L, 0L, 1L, 1L, 1L, 
0L, 0L, 0L, 0L, 1L, 0L, 1L, 0L, 1L, 1L, 0L, 0L, 0L, 0L, 1L, 1L, 
1L, 0L, 0L, 0L, 0L, 1L, 1L, 0L, 1L, 0L, 1L, 1L, 0L, 1L, 0L, 0L, 
0L, 0L, 1L, 0L, 1L, 1L, 0L, 1L, 0L, 0L, 1L, 0L, 0L, 0L, 0L, 1L, 
0L, 0L, 0L, 0L, 1L, 0L, 1L, 0L, 0L, 0L, 1L, 0L, 0L, 1L, 0L, 1L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 1L, 0L, 1L, 1L, 1L, 1L, 1L, 0L, 1L, 
0L, 0L, 1L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 1L, 0L, 1L, 0L, 1L, 0L, 
1L, 1L, 0L, 0L, 0L, 0L, 1L, 1L, 1L, 0L, 0L, 0L, 0L, 0L, 0L, 1L, 
1L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 1L, 0L, 1L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 1L, 1L, 0L, 0L, 0L, 0L, 1L, 1L, 1L, 
1L, 0L, 1L, 1L, 0L, 1L, 0L, 1L, 1L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
1L, 1L, 0L, 1L, 1L, 1L, 1L, 0L, 1L, 0L, 1L, 0L, 1L, 0L, 0L, 1L, 
1L, 0L, 0L, 0L, 0L, 0L, 1L, 1L, 0L, 0L, 0L, 0L, 1L, 1L, 0L, 1L, 
0L, 1L, 1L, 1L, 0L, 1L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 1L, 1L, 0L, 
0L, 1L, 0L, 1L, 1L, 1L, 1L, 0L, 0L, 0L, 0L, 1L, 0L, 0L, 0L, 0L, 
1L, 0L, 0L, 0L, 0L, 0L, 1L, 0L, 0L, 1L, 0L, 1L, 0L, 0L, 0L, 1L, 
0L, 0L, 1L, 0L, 0L, 1L, 0L, 0L, 1L, 1L, 0L, 1L, 1L, 0L, 0L, 1L, 
0L, 0L, 1L, 1L, 0L, 0L, 0L, 1L, 0L, 1L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 1L, 0L, 0L, 0L, 0L, 0L, 1L, 1L, 0L, 1L, 1L, 0L, 0L, 0L, 
1L, 0L, 1L, 0L, 0L, 0L, 0L, 0L, 1L, 0L, 1L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 1L, 0L, 0L, 1L, 0L, 0L, 0L, 1L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 1L, 1L, 0L, 0L, 0L, 0L, 0L, 1L, 0L, 0L, 1L, 0L, 1L, 0L, 
0L, 0L, 0L, 1L, 0L, 0L, 0L, 0L, 1L, 0L, 1L, 0L, 0L, 0L, 0L, 0L, 
1L, 0L, 0L, 0L, 0L, 0L, 0L, 1L, 1L, 1L, 0L, 1L, 0L, 0L, 0L, 0L, 
1L, 1L, 1L, 1L, 0L, 0L, 1L, 0L, 0L, 0L, 0L, 0L, 1L, 0L, 0L, 0L, 
0L, 0L, 0L, 1L, 0L, 0L, 1L, 0L, 0L, 0L, 0L, 0L, 1L, 0L, 0L, 0L, 
0L, 0L, 0L, 1L, 1L, 1L, 1L, 0L, 1L, 0L, 0L, 0L, 0L, 0L, 0L, 1L, 
0L, 0L, 1L, 0L, 1L, 1L, 1L, 1L, 1L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
1L, 1L, 1L, 1L, 0L, 0L, 1L, 0L, 0L, 0L, 1L, 1L, 0L, 0L, 0L, 0L, 
1L, 1L, 1L, 1L, 1L, 1L, 1L, 0L, 1L, 1L, 0L, 1L, 1L, 0L, 0L, 1L, 
1L, 0L, 1L, 0L, 1L, 1L, 0L, 1L, 1L, 1L, 1L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 1L, 1L, 1L, 1L, 0L, 0L, 0L, 0L, 0L, 1L, 0L, 1L, 0L, 0L, 
1L, 1L, 1L, 0L, 0L, 0L, 0L, 0L, 1L, 0L, 1L, 0L, 0L, 0L, 1L, 1L, 
1L, 1L, 0L, 0L, 1L, 0L, 1L, 0L, 0L, 0L, 0L, 0L, 1L, 1L, 0L, 1L, 
0L, 0L, 1L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 1L, 1L, 
0L, 0L, 1L, 0L, 1L, 0L, 0L, 1L, 1L, 1L, 1L, 0L, 1L, 0L, 0L, 0L, 
1L, 1L, 0L, 0L, 0L, 1L, 1L, 0L, 0L, 0L, 1L, 0L, 0L, 0L, 0L, 0L, 
1L, 0L, 1L, 0L, 0L, 0L, 1L, 0L, 1L, 1L, 1L, 0L, 0L, 1L, 1L, 1L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 1L, 0L, 0L, 0L, 0L, 1L, 0L, 0L, 0L, 
0L, 1L, 0L, 0L, 0L, 0L, 1L, 0L, 0L, 1L, 0L, 1L, 0L, 1L, 1L, 0L, 
0L, 0L, 1L, 1L, 0L, 0L, 0L, 1L, 0L, 1L, 1L, 0L, 0L, 0L, 0L, 1L, 
0L, 0L, 0L, 0L, 1L, 0L, 0L, 0L, 0L, 0L), Race = c(2L, 1L, 1L, 
3L, 2L, 2L, 3L, 2L, 2L, 1L, 3L, 1L, 1L, 2L, 4L, 3L, 2L, 3L, 3L, 
3L, 1L, 5L, 2L, 2L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 4L, 3L, 2L, 1L, 
1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 3L, 1L, 1L, 4L, 1L, 1L, 3L, 1L, 
1L, 1L, 3L, 3L, 3L, 1L, 3L, 5L, 2L, 3L, 1L, 1L, 3L, 1L, 2L, 1L, 
2L, 1L, 2L, 2L, 2L, 2L, 4L, 3L, 3L, 2L, 3L, 1L, 2L, 3L, 4L, 2L, 
1L, 2L, 2L, 2L, 3L, 3L, 4L, 4L, 3L, 4L, 5L, 2L, 2L, 1L, 3L, 3L, 
3L, 4L, 1L, 2L, 3L, 1L, 3L, 2L, 1L, 1L, 1L, 3L, 3L, 1L, 3L, 3L, 
3L, 1L, 3L, 1L, 1L, 1L, 1L, 2L, 3L, 2L, 3L, 3L, 1L, 2L, 3L, 5L, 
4L, 1L, 3L, 2L, 3L, 5L, 2L, 3L, 1L, 3L, 1L, 4L, 2L, 5L, 3L, 2L, 
1L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 1L, 3L, 2L, 2L, 3L, 3L, 1L, 
3L, 3L, 3L, 1L, 3L, 3L, 3L, 3L, 3L, 1L, 1L, 4L, 1L, 3L, 2L, 2L, 
1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 5L, 1L, 1L, 3L, 4L, 5L, 5L, 
1L, 1L, 3L, 4L, 1L, 4L, 1L, 3L, 1L, 3L, 1L, 1L, 4L, 3L, 2L, 1L, 
4L, 5L, 3L, 3L, 1L, 1L, 4L, 2L, 1L, 3L, 3L, 1L, 1L, 1L, 2L, 2L, 
3L, 1L, 1L, 1L, 2L, 1L, 3L, 3L, 1L, 3L, 3L, 1L, 1L, 3L, 1L, 3L, 
1L, 1L, 1L, 1L, 1L, 2L, 3L, 1L, 1L, 1L, 1L, 1L, 1L, 2L, 3L, 3L, 
1L, 1L, 2L, 1L, 2L, 1L, 5L, 2L, 2L, 2L, 1L, 1L, 2L, 1L, 1L, 2L, 
2L, 1L, 1L, 5L, 1L, 1L, 5L, 2L, 3L, 1L, 4L, 4L, 4L, 2L, 2L, 2L, 
3L, 1L, 5L, 5L, 1L, 5L, 5L, 5L, 5L, 2L, 5L, 1L, 5L, 1L, 3L, 3L, 
1L, 2L, 2L, 1L, 1L, 2L, 1L, 2L, 3L, 4L, 2L, 1L, 1L, 1L, 2L, 2L, 
3L, 2L, 3L, 2L, 4L, 2L, 1L, 3L, 1L, 3L, 1L, 1L, 1L, 2L, 1L, 1L, 
1L, 1L, 1L, 4L, 1L, 1L, 1L, 1L, 5L, 3L, 3L, 3L, 3L, 3L, 3L, 5L, 
4L, 3L, 3L, 3L, 1L, 3L, 3L, 3L, 1L, 4L, 4L, 3L, 5L, 3L, 2L, 3L, 
5L, 2L, 4L, 4L, 3L, 4L, 4L, 2L, 3L, 2L, 1L, 2L, 4L, 3L, 1L, 3L, 
3L, 3L, 5L, 2L, 5L, 5L, 3L, 2L, 3L, 3L, 1L, 3L, 4L, 2L, 2L, 3L, 
2L, 3L, 1L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 
3L, 2L, 2L, 3L, 3L, 3L, 3L, 3L, 3L, 2L, 2L, 3L, 3L, 2L, 3L, 3L, 
3L, 1L, 1L, 2L, 3L, 1L, 5L, 3L, 3L, 2L, 2L, 3L, 3L, 2L, 2L, 3L, 
3L, 1L, 1L, 3L, 3L, 3L, 1L, 3L, 1L, 1L, 1L, 1L, 2L, 1L, 1L, 1L, 
1L, 1L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 3L, 3L, 2L, 2L, 
2L, 3L, 2L, 3L, 3L, 3L, 3L, 2L, 2L, 4L, 2L, 3L, 2L, 3L, 3L, 2L, 
1L, 2L, 1L, 1L, 2L, 2L, 3L, 2L, 2L, 2L, 3L, 2L, 2L, 2L, 2L, 2L, 
3L, 3L, 2L, 3L, 3L, 3L, 1L, 1L, 2L, 2L, 1L, 1L, 4L, 1L, 3L, 3L, 
3L, 2L, 1L, 2L, 3L, 3L, 3L, 2L, 2L, 2L, 3L, 3L, 3L, 3L, 2L, 2L, 
2L, 3L, 2L, 2L, 4L, 5L, 1L, 5L, 1L, 1L, 5L, 5L, 3L, 3L, 3L, 3L, 
1L, 2L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 1L, 3L, 3L, 
3L, 3L, 3L, 1L, 2L, 3L, 3L, 1L, 1L, 3L, 1L, 5L, 1L, 2L, 1L, 3L, 
4L, 1L, 2L, 5L, 2L, 2L, 2L, 2L, 3L, 1L, 3L, 3L, 3L, 1L, 1L, 2L, 
1L, 1L, 2L, 2L, 2L, 2L, 3L, 2L, 3L, 2L, 2L, 2L, 2L, 1L, 2L, 2L, 
2L, 2L, 3L, 1L, 2L, 3L, 2L, 3L, 1L, 3L, 2L, 3L, 3L, 4L, 1L, 1L, 
3L, 3L, 2L, 3L, 1L, 3L, 4L, 3L, 3L, 1L, 3L, 2L, 3L, 4L, 3L, 2L, 
2L, 2L, 1L, 3L, 3L, 2L, 1L, 3L, 3L, 3L, 2L, 3L, 2L, 3L, 3L, 3L, 
1L, 3L, 3L, 1L, 3L, 1L, 4L, 4L, 1L, 3L, 4L, 3L, 4L, 2L, 2L, 1L, 
2L, 4L, 3L, 2L, 3L, 3L, 2L, 4L, 5L, 1L, 4L, 4L, 3L, 4L, 1L, 3L, 
3L, 3L, 2L, 3L, 3L, 2L, 3L, 3L, 2L, 3L, 3L, 3L, 2L, 1L, 1L, 3L, 
3L, 2L, 2L, 2L, 3L, 3L, 2L, 3L, 1L, 2L, 2L, 3L, 1L, 4L, 3L, 3L, 
3L, 3L, 2L, 2L, 1L, 3L, 3L, 2L, 3L, 1L, 3L, 3L, 2L, 2L, 1L, 1L, 
2L, 1L, 2L, 2L, 3L, 1L, 2L, 2L, 1L, 2L, 2L, 2L, 1L, 2L, 2L, 3L, 
3L, 2L, 3L, 5L, 4L, 3L, 2L, 2L, 5L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
1L, 2L), BedSize = c(4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 
4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 2L, 2L, 2L, 2L, 2L, 2L, 
2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 
3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 2L, 2L, 2L, 2L, 2L, 2L, 4L, 
4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 
4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 3L, 3L, 3L, 2L, 
2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 3L, 3L, 3L, 3L, 3L, 3L, 
3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 
3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 2L, 2L, 2L, 2L, 2L, 2L, 
2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 
1L, 1L, 1L, 1L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 
3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 2L, 2L, 
2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 3L, 3L, 3L, 3L, 
3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 
3L, 3L, 3L, 3L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
2L, 2L, 2L, 2L, 2L, 2L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 
3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 
3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 2L, 2L, 
2L, 2L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 
3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 
2L, 2L, 2L, 2L, 2L, 3L, 3L, 3L, 3L, 3L, 3L, 2L, 2L, 2L, 2L, 4L, 
4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 
4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 
4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 2L, 2L, 2L, 2L, 2L, 2L, 
2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 3L, 3L, 3L, 
3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 2L, 2L, 2L, 2L, 2L, 
2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 4L, 3L, 3L, 3L, 3L, 3L, 3L, 
3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 
3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 2L, 2L, 2L, 2L, 
2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 3L, 3L, 3L, 
1L, 1L, 1L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
2L, 3L, 3L, 3L, 3L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
2L, 2L, 2L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 
3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 
3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 2L, 2L, 2L, 2L, 2L, 2L, 
2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 4L, 4L, 4L, 4L, 4L, 
4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 4L, 
3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 
3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 3L, 
3L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L), Owner = c(2L, 2L, 
2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 1L, 
1L, 1L, 1L, 1L, 1L, 1L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
2L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 
1L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 1L, 1L, 
2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
2L, 2L, 2L, 2L, 2L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 
1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 
1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 
1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 
1L, 1L, 1L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 1L, 1L, 1L, 1L, 1L, 1L, 
1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 
1L, 1L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
2L, 2L, 2L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 
1L, 1L, 1L, 1L, 1L, 1L, 1L, 1L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 2L, 
2L, 2L, 2L), Compl = c(0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 1L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 1L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 1L, 0L, 0L, 
0L, 0L, 0L, 1L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 1L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 1L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 1L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 1L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 1L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 1L, 0L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 1L, 0L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 1L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 1L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 1L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 1L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 1L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 
0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L, 0L), Staff = c(4.6419, 
4.6419, 4.6419, 4.6419, 4.6419, 4.6419, 4.6419, 4.6419, 4.6419, 
4.6419, 4.6419, 4.6419, 4.6419, 4.6419, 4.6419, 4.6419, 4.6419, 
8.6794, 8.6794, 8.6794, 4.3699, 4.3699, 4.3699, 4.3699, 4.3363, 
4.3363, 4.3363, 4.3363, 4.3363, 4.3363, 4.3363, 4.3363, 4.3363, 
4.3363, 4.3363, 4.3363, 4.3363, 4.3363, 4.3363, 4.3363, 4.3363, 
4.3363, 4.3363, 4.3363, 4.3363, 4.3363, 4.3363, 4.3363, 4.3363, 
4.3363, 4.3363, 4.7452, 4.7452, 4.7452, 4.7452, 4.7452, 4.7452, 
4.7452, 4.7452, 4.7452, 4.7452, 4.7452, 4.7452, 4.7452, 4.7452, 
4.7452, 4.7452, 3.8462, 3.8462, 3.8462, 3.8462, 3.8462, 3.8462, 
6.4841, 6.4841, 6.4841, 6.4841, 6.4841, 6.4841, 6.4841, 6.4841, 
6.4841, 6.4841, 6.4841, 6.4841, 6.4841, 6.4841, 6.4841, 6.4841, 
6.4841, 6.4841, 6.4841, 6.4841, 6.4841, 6.4841, 6.4841, 6.4841, 
6.4841, 6.4841, 6.4841, 6.4841, 6.4841, 9.3671, 9.3671, 9.3671, 
2.8125, 2.8125, 2.8125, 2.8125, 2.8125, 2.8125, 2.8125, 2.8125, 
2.8125, 2.8125, 2.8125, 3.6825, 3.6825, 3.6825, 3.6825, 3.6825, 
3.6825, 3.6825, 3.6825, 3.6825, 3.6825, 3.6825, 3.6825, 3.6825, 
3.6825, 3.6825, 3.6825, 3.6825, 3.6825, 3.6825, 3.6825, 3.6825, 
3.6825, 3.6825, 3.6825, 3.6825, 3.6825, 3.6825, 3.6825, 3.6825, 
3.6825, 3.6825, 3.6825, 3.4614, 3.4614, 3.4614, 3.4614, 3.4614, 
3.4614, 3.4614, 3.4614, 3.4614, 3.4614, 3.4614, 3.4614, 3.4614, 
3.4614, 3.4614, 3.4614, 3.4614, 3.4614, 3.4614, 3.4614, 3.4614, 
3.4614, 3.4614, 3.4614, 3.4614, 3.4614, 3.4614, 3.4614, 3.4614, 
3.4614, 3.4614, 3.9658, 3.9658, 3.9658, 3.9658, 3.9658, 3.9658, 
3.9658, 3.9658, 3.9658, 3.9658, 3.9658, 6.9006, 6.9006, 6.9006, 
6.9006, 6.9006, 6.9006, 6.9006, 6.9006, 6.9006, 6.9006, 6.9006, 
6.9006, 3.5233, 3.5233, 3.5233, 3.5233, 3.5233, 3.5233, 3.5233, 
3.5233, 3.5233, 3.5233, 3.5233, 3.5233, 3.5233, 3.5233, 3.5233, 
3.5233, 3.5233, 3.5233, 3.5233, 3.5233, 3.5233, 3.5233, 3.5233, 
3.5233, 3.5233, 3.5233, 3.5233, 3.5233, 3.5233, 3.5233, 7.4683, 
7.4683, 7.4683, 7.4683, 7.4683, 7.4683, 7.4683, 9.5487, 9.5487, 
9.5487, 9.5487, 9.5487, 9.5487, 9.5487, 4.5611, 4.5611, 4.5611, 
4.5611, 4.5611, 4.5611, 4.5611, 4.5611, 4.5611, 4.5611, 3.1863, 
4.7026, 4.7026, 4.7026, 4.7026, 4.7026, 4.7026, 4.7026, 4.7026, 
4.7026, 4.7026, 4.7026, 4.7026, 4.7026, 8.7974, 8.7974, 8.7974, 
8.7974, 8.7974, 8.7974, 8.7974, 8.7974, 8.7974, 8.7974, 8.7974, 
8.7974, 8.7974, 4.1099, 4.1099, 4.1099, 4.1099, 4.1099, 4.1099, 
4.1099, 4.1099, 4.1099, 4.1099, 5.5977, 5.5977, 5.5977, 5.5977, 
5.5977, 5.5977, 5.5977, 5.5977, 5.5977, 5.5977, 5.5977, 4.7452, 
4.7452, 4.1043, 4.1043, 4.1043, 4.1043, 4.1043, 4.1043, 3.2559, 
3.2559, 3.2559, 3.2559, 3.2559, 3.2559, 3.2559, 3.2559, 3.2559, 
3.2559, 3.2559, 3.2559, 3.2559, 3.2559, 3.2559, 3.2559, 3.2559, 
3.2559, 3.2559, 3.2559, 3.2559, 3.2559, 3.2559, 3.2559, 3.2559, 
3.2559, 2.5606, 2.5606, 2.5606, 2.5606, 2.5606, 2.5606, 9.5487, 
9.5487, 9.5487, 9.5487, 3.5175, 3.5175, 3.5175, 3.5175, 3.5175, 
3.5175, 3.5175, 3.5175, 3.5175, 3.5175, 3.5175, 3.5175, 3.5175, 
3.5175, 3.5175, 3.5175, 3.5175, 3.5175, 3.5175, 3.5175, 3.5175, 
3.5175, 3.5175, 3.5175, 3.5175, 3.5175, 3.5175, 3.5175, 3.5175, 
3.5175, 3.8462, 3.8462, 3.8462, 3.8462, 3.8462, 4.1043, 4.1043, 
4.1043, 3.5175, 3.5175, 3.5175, 2.9617, 2.9617, 2.9617, 2.9617, 
4.6419, 4.6419, 4.6419, 4.6419, 4.6419, 4.6419, 4.6419, 4.6419, 
4.6419, 4.6419, 4.6419, 4.6419, 4.6419, 4.6419, 8.6794, 8.6794, 
8.6794, 8.6794, 8.6794, 8.6794, 8.6794, 8.6794, 8.6794, 8.6794, 
8.6794, 8.6794, 8.6794, 8.6794, 8.6794, 8.6794, 8.6794, 8.6794, 
8.6794, 8.6794, 8.6794, 8.6794, 8.6794, 8.6794, 8.6794, 8.6794, 
8.6794, 8.6794, 8.6794, 4.3699, 4.3699, 4.3699, 4.3699, 4.3699, 
4.3699, 4.3699, 4.3699, 4.3699, 4.3699, 4.3699, 4.3699, 4.3699, 
4.3699, 4.3699, 4.3699, 4.3699, 4.3699, 4.3699, 4.7452, 4.7452, 
4.7452, 4.7452, 4.7452, 4.7452, 4.7452, 4.7452, 4.7452, 4.7452, 
4.7452, 4.7452, 4.7452, 4.7452, 3.8462, 3.8462, 3.8462, 3.8462, 
3.8462, 3.8462, 3.8462, 3.8462, 3.8462, 3.8462, 3.8462, 3.8462, 
3.8462, 3.8462, 6.4841, 9.3671, 9.3671, 9.3671, 9.3671, 9.3671, 
9.3671, 9.3671, 9.3671, 9.3671, 9.3671, 9.3671, 9.3671, 9.3671, 
9.3671, 9.3671, 9.3671, 9.3671, 9.3671, 9.3671, 9.3671, 9.3671, 
9.3671, 9.3671, 9.3671, 9.3671, 9.3671, 9.3671, 9.3671, 9.3671, 
9.3671, 9.3671, 9.3671, 9.3671, 9.3671, 2.8125, 2.8125, 2.8125, 
2.8125, 2.8125, 2.8125, 2.8125, 2.8125, 2.8125, 2.8125, 2.8125, 
2.8125, 2.8125, 2.8125, 2.8125, 2.8125, 2.8125, 3.6825, 3.6825, 
3.6825, 3.9658, 3.9658, 3.9658, 7.4683, 7.4683, 7.4683, 7.4683, 
6.9006, 6.9006, 6.9006, 6.9006, 6.9006, 6.9006, 6.9006, 6.9006, 
6.9006, 6.9006, 3.5233, 3.5233, 3.5233, 3.5233, 7.4683, 7.4683, 
7.4683, 7.4683, 7.4683, 7.4683, 7.4683, 7.4683, 7.4683, 7.4683, 
7.4683, 7.4683, 7.4683, 7.4683, 7.4683, 7.4683, 7.4683, 7.4683, 
7.4683, 7.4683, 7.4683, 7.4683, 7.4683, 9.5487, 9.5487, 9.5487, 
9.5487, 9.5487, 9.5487, 9.5487, 4.5611, 4.5611, 4.5611, 4.5611, 
4.5611, 4.5611, 4.5611, 4.5611, 4.5611, 3.1863, 3.1863, 3.1863, 
3.1863, 3.1863, 3.1863, 3.1863, 3.1863, 3.1863, 3.1863, 3.1863, 
3.1863, 3.1863, 3.1863, 3.1863, 4.7026, 4.7026, 4.7026, 4.7026, 
4.7026, 4.7026, 4.7026, 4.7026, 4.7026, 4.7026, 4.7026, 4.7026, 
4.7026, 4.7026, 4.7026, 8.7974, 8.7974, 8.7974, 8.7974, 8.7974, 
8.7974, 8.7974, 8.7974, 8.7974, 8.7974, 8.7974, 8.7974, 8.7974, 
8.7974, 8.7974, 8.7974, 8.7974, 8.7974, 8.7974, 8.7974, 8.7974, 
8.7974, 8.7974, 8.7974, 6.2894, 6.2894, 6.2894, 6.2894, 6.2894, 
6.2894, 6.2894, 6.2894, 6.2894, 6.2894, 6.2894, 6.2894, 6.2894, 
6.2894, 6.2894, 6.2894, 6.2894, 6.2894, 6.2894, 6.2894, 6.2894, 
6.2894, 6.2894, 6.2894, 4.1099, 4.1099, 4.1099, 4.1099, 4.1099, 
4.1099, 4.1099, 4.1099, 4.1099, 4.1099, 4.1099, 4.1099, 4.1099, 
4.1099, 4.1099, 4.1099, 4.1099, 4.1099, 4.1099, 4.1099, 4.1099, 
4.1099, 5.5977, 5.5977, 5.5977, 5.5977, 5.5977, 5.5977, 5.5977, 
5.5977, 5.5977, 5.5977, 5.5977, 10.766, 10.766, 10.766, 10.766, 
10.766, 10.766, 10.766, 10.766, 10.766, 10.766, 10.766, 10.766, 
10.766, 10.766, 10.766, 10.766, 10.766, 10.766, 10.766, 10.766, 
10.766, 4.1043, 4.1043, 4.1043, 4.1043, 4.1043, 4.1043, 4.1043, 
4.1043, 4.1043, 4.1043, 4.1043, 4.1043, 4.1043, 4.1043, 4.1043, 
4.1043, 4.1043, 4.1043, 4.1043, 4.1043, 4.1043, 4.1043, 4.1043, 
3.2559, 3.2559, 2.5606, 2.5606, 2.5606, 2.5606, 2.5606, 3.2559, 
3.2559, 3.2559, 3.8462, 3.8462, 3.8462, 3.8462, 3.8462, 4.1099, 
4.1099, 4.1099, 4.1099, 4.1099, 4.1099, 4.1099, 4.1099, 4.1099, 
6.9006, 6.9006, 6.9006, 6.9006, 6.9006, 6.9006, 6.9006, 6.9006, 
6.9006, 6.9006, 6.9006, 6.9006)), .Names = c("LOS", "Hospital", 
"Insurer", "Age", "Gender", "Race", "BedSize", "Owner", "Compl", 
"Staff"), class = "data.frame", row.names = c(NA, -789L))

par( mfrow=c(1,2))

pie( table(factor(los$Race)), 
	labels=c("White","Hispanic","Black","Asian","Other"), 
   col=c("white","olivedrab","black","yellow","brown"),
   main="Piechart of number of children\nhospitalized with asthma by race")
   
plot( factor(los$Race), 
	las=2, 
	col="slateblue", 
	names=c("White","Hispanic","Black","Asian","Other"), 
	ylim=c(0, 300),
   sub="Race", 
   ylab="Number of children", 
   main="Barchart of number of children\nhospitalized with asthma by race")

par( mfrow=c(1,1))
```

   

```{exercise}
Rank the categories from largest to smallest
using each graph in
Fig. \@ref(fig:3Dcharts).
Which graph makes it easiest?
```

```{r 3Dcharts, echo=FALSE, fig.cap="Three different graphs for the same data"}
# New data
library(plotrix)

par(mfrow=c(2,2), mar=c(1, 1, 1, 1))

set.seed(222)
nos <- round(40 + rnorm(5,0,10))
nos <- c(32, 36, 37, 30, 34)
labs <- c("A", "B", "C", "D", "E")

par( mar=c(2,2,2,2))
pie3D( nos, theta=pi/4, 
   labels=labs)

pie( nos, 
   labels=labs)
   
#barplot(nos, las=1, names.arg=labs, ylab="Number")
#abline(h=seq(0,35,by=5), col="white")

barplot(nos, 
   las=1, 
   ylab="Number", 
   axes=FALSE)
axis(side=1, labels=labs, at=1:5 )

abline(h=seq(0,35,by=5), col="white")

par(mfrow=c(1,1))
```

Recall that the purpose of a graph is to 
is to display the information 
in the clearest, simplest possible way,
to help us understand the data.
Usually,
adding an artificial third dimension does not make the message clearer.



```{exercise}
Consider an extreme example of the distortion
created when using three-dimensions
(Fig. \@ref(fig:Extreme3D)).
Estimate the percentage of each pie slice.
```


```{r Extreme3D, echo=FALSE, fig.cap="Extreme 3D pie charts"}
par( mar=c(0.25, 0.25, 0.25, 0.25))

pie3D( c(30, 20, 20, 30), 
       labels=c("A", "B","C", "D"),
       theta=pi/20,  
       start=2*pi/3)
```



### Do we even need a graph?

 
Although graphs are excellent,
sometimes a graphic is not the best way to display information.
Sometimes just writing the information down is better
('14.29\% of the patients were suitable for the study'),
as above).
Sometimes a table may be better
for example,
when there are a small number of levels,
or if more details might be useful.
Compare these two ways of presenting some data:
Fig. \@ref(fig:SimpleBar)
and
Table \@ref(tab:SimpleTable)







```{r SimpleBar, echo=FALSE, fig.cap="Simple data, displayed as a barchart"}

# New data
library(plotrix)

par(mar=c(2, 2, 0, 0))
set.seed(222)
nos <- round(40 + rnorm(5,0,10))
nos <- c(32, 36, 37, 30, 34)
labs <- c("A", "B", "C", "D", "E")


barplot(nos, 
   las=1, 
   ylab="Number", 
   axes=FALSE)
axis(side=1, labels=labs, at=1:5 )

abline(h=seq(0,35,by=5), col="white")

par(mfrow=c(1,1))
```



```{r SimpleTable, echo=FALSE}
simpletab <- data.frame(Response=labs,
                        Numbers=nos, 
                        Percentage=round(nos/sum(nos)*100, 0) )


kable(simpletab,
      align=c("r", "r", "r"),
      booktabs = TRUE,
      caption="Simple data, displayed as a table")
```


	
      













## One qualitative variable and one quantitative variable


Relational and interventional RQs are about the relationships between variables.
These relationships,
and other relationships between two variables,
can also be examined by using suitable graphs.


With one qualitative variable and one quantitative variable,
suitable graphs include: 

* *Back-to-back stem-and-leaf plot*:
         Good choice for small amounts of data
         when the qualitative variable only has *two levels*; 
* *2-D Dot chart*: 
         Good choice for small amounts of data;
* *Boxplot*:
         Good choice, except for small amounts of data.        
         










### Back-to-back stem-and-leaf

Back-to-back stem-and-leaf plots
are essentially two stem-and-leaf plots that share the stems;
one group has the leaves going left-to-right from the stem,
and 
a second group has the leaves going right-to-left from the stem.


```{example}
A study of krill [@data:Greenacre2016:reporting]
produced this back-to-back stem-and-leaf plot:

We can see that the treatment group tends to produce more eggs.
```
			
		
		
### 2-D Dot charts


A 2-D dot charts places dots for each observation,
but separated for each level of the qualitative variable.

```{example}
In a study of Vitamin D [@data:Hall2010:vitaminD],
the following dot chart was produced:

MAKE IMAGE
```








### Boxplots


Understanding boxplots takes some explanation,
and we will return to boxplots again later (XREF!).


Let's try to explain how a boxplot works.
To do so,
we'll focus on just one single boxplot from one variable
(though it is best to use histograms for one quantitative variable).
The boxplot for Age in the NHANES data is shown in
Fig.~\ref{FG:GRAPHICS:NHANESAgeBoxplot} (left panel). 

NHANES DATA HASNT BEEN INTRODUCED!


```{r echo=FALSE}
library(NHANES)

Age.quantiles <- quantile(NHANES$Age)
#  0%  25%  50%  75% 100% 
#   0   18   36   54   80 
HDL.quantiles <- quantile(NHANES$DirectChol, na.rm=TRUE)
#  0%  25%  50%  75% 100% 
#0.52 1.06 1.29 1.58 3.44 
```

The boxplot essentially has five horizontal lines:

* The oldest subject is about `r Age.quantiles[5]`:
This is the line at the top of the boxplot.
*  75\% of subjects are younger than about `r Age.quantiles[4]`,
   and this is represented by the line at the top of the 
   central box.
* 50\% of subjects are younger than about `r Age.quantiles[3]`,
   and this is represented by the line in the centre of the 
   central box.
* 25\% of subjects are younger than about `r Age.quantiles[2]`,
and this is represented by the line at the bottom of the 
central box.
* The youngest subject is about `r Age.quantiles[1]`.
   This is the line at the bottom of the boxplot.


```{r echo=FALSE}
par(mfrow=c(1,2))

par(mar=c(0,5,0,0) + 0.5)
boxplot(NHANES$Age, 
	col="wheat", 
	las=1,
	ylim=c(0, 80),
	ylab="Age (in years)"
)

boxplot(NHANES$DirectChol, 
	col="wheat", 
	las=1,
	ylim=c(0, 4),
	ylab="Direct HDL cholesterol (mmol/L)"
)

par(mfrow=c(1,1))
```



A variation of the boxplot is shown in
Fig.~\ref{FG:GRAPHICS:NHANESAgeBoxplot} (right panel). 
The boxplot is produced using the same principles as the boxplot in the left panel,
except that
the horizontal line extending from the central box
extends to the 'maximum' (except for outliers),
and 
the horizontal line extending from the central box
extends to the 'minimum' (except for outliers).
That is,
the lines extend from the box as far as possible,
*except* for outliers (unusual observations).
We will revisit boxplots in the next chapter,
and explain how SPSS decides which observations are outliers
(XREF!).


```{example}
A study [@data:Ondei2017:monsoon] compared various rainforests on various
criteria, 
including the moisture index.
The following boxplot compares rainforests in Western Australia and the Northern Territory:
```


```{example}
A study of different engineering project delivery methods [@data:Hale2009:ProjectDelivery]
produced the boxplot below:
```
	



```{example}
Boxplots can be horizontal too,
which leaves more room for long labels;
for example [@data:Silva2011:StoneCladding]:
```


		
		


## Two quantitative variables {#Scatterplots}

Scatterplots display the relationship between *two quantitative variables*.
Conventionally,
the "response" variable is on the vertical axis, and the
the "explanatory" variable is on the horizontal axis.
As with any graph,
we need to explain what the graph tells us.
The purpose of a graph is to display the information 
in the clearest, simplest possible way,
to help us understand the data.
For a *scatterplot*,
we can describe briefly how the variables are related to each other.


```{example}
A study of atheltes at the Australian Institute of Sport (AIS)
measured numerous physical and blood measurements from high performance athletes
[@data:Telford1991:sexsportsize].
Many relationships were of interest
Fig. \@ref(fig:ScatterAIS)
shows the relationship between the sum of skin folds and percentage body fat.
Each point represents one athlete.
Clearly,
the greater the percentage body fat,
the greater the sum of skin folds, in general.  
```
	
```{r ScatterAIS, echo=FALSE, results ='hide', fig.cap="Scatter plot of SSF against percentage body fat for male AIS athletes"}
### TAKEN FROM GLMsData:  dput( subset(AIS, Sex=="M")[, c("SSF", "PBF")])
library(GLMsData)
data(AIS)

AIS.sub <- dput( subset(AIS, Sex=="M")[, c("SSF", "PBF")])
structure(list(SSF = c(44.5, 41.8, 33.7, 50.9, 40.5, 51.2, 54.4, 
52.3, 57, 65.3, 52, 42.7, 35.2, 49.2, 61.8, 46.5, 34.8, 60.2, 
48.1, 44.5, 54, 44.7, 64.9, 43.8, 58.3, 52.8, 43.1, 78, 40.8, 
41.5, 50.9, 49.6, 88.9, 48.3, 61.8, 43, 61.1, 43.8, 54.2, 41.8, 
34.1, 30.5, 34, 46.7, 71.1, 65.9, 34.3, 34.6, 31.8, 34.5, 31, 
32.6, 31.5, 32.6, 31, 28, 33.7, 30.3, 38, 55.7, 37.5, 112.5, 
82.7, 29.7, 38.9, 44.8, 30.9, 44, 37.5, 37.6, 31.7, 36.6, 48, 
41.9, 30.9, 52.8, 43.2, 113.5, 96.9, 49.3, 42.3, 96.3, 56.5, 
105.7, 100.7, 56.8, 75.9, 52.8, 47.8, 76, 61.2, 75.6, 43.3, 49.5, 
70, 75.7, 57.7, 67.2, 56.5, 47.6, 60.4, 34.9), PBF = c(8.47, 
7.68, 6.16, 8.56, 6.86, 9.4, 9.17, 8.54, 9.2, 11.72, 8.44, 7.19, 
6.46, 9, 12.61, 9.03, 6.96, 10.05, 9.56, 9.36, 10.81, 8.61, 9.53, 
7.42, 9.79, 8.97, 7.49, 11.95, 7.35, 7.16, 8.77, 9.56, 14.53, 
8.51, 10.64, 7.06, 8.87, 7.88, 9.2, 7.19, 6.06, 5.63, 6.59, 9.5, 
13.97, 11.66, 6.43, 6.99, 6, 6.56, 6.03, 6.33, 6.82, 6.2, 5.93, 
5.8, 6.56, 6.76, 7.22, 8.51, 7.72, 19.94, 13.91, 6.1, 7.52, 9.56, 
6.06, 7.35, 6, 6.92, 6.33, 5.9, 8.84, 8.94, 6.53, 9.4, 8.18, 
17.41, 18.08, 9.86, 7.29, 18.72, 10.12, 19.17, 17.24, 9.89, 13.06, 
8.84, 8.87, 14.69, 8.64, 14.98, 7.82, 8.97, 11.63, 13.49, 10.25, 
11.79, 10.05, 8.51, 11.5, 6.26)), .Names = c("SSF", "PBF"), class = "data.frame", row.names = 101:202)

plot( SSF ~ PBF,
      data=AIS.sub,
      las=1,
      xlim=c(5, 20),
      ylim=c(20, 120),
      main="SSF vs PBF: Male AIS athletes",
      xlab="Percentage body fat",
      ylab="Sum of skin fold",
      sub="(Note: Units of measurement for skin folds not given)",
      pch=19)
```






## Two qualitative variables

The relationship between two qualitative variables
can be explored using:

* *Side-by-side bar charts*;
* *Stacked bar charts*; or
* *Dot charts*.

Sometimes the best summary is just a two-way
table of counts (or percentages).
      


```{example}
```{r echo=FALSE}
## Russell2009

Counts <- matrix( c(75, 25, 21,   74, 27, 22,   71, 10, 18,   58, 10, 12), nrow=4, ncol=3, byrow=TRUE) 
rownames(Counts) <- c("Autumn", "Winter", "Spring", "Summer")
colnames(Counts) <- c("Unknown", "M", "F")
```

A study of roadkill [@data:Russell2009:Possums] produced this stacked bar chart:


Many other variations of this chart were possible:

* The bars could have been horizontal rather than vertical;
* Percentages, rather than counts, could have been used;
* Each bar could represent the sex of the possum,
		and the seasons stacked;
* The section within each bart could have been placed side-by-side,
		rather than stacked.

That is,
there are many variations of the this graph;
some are shown in 
Fig. \@ref(fig:PossumBarCharts).
	

```{r PossumBarCharts, echo=FALSE, fig.height=14, fig.width=5, fig.cap="Four graphs: The number of possums found as road kill, by sex and season"}
par( mfrow=c(4, 1))
colnames(Counts) <- c("?", "M", "F")

barplot( t(Counts), 
   beside=TRUE,
   ylab="Number of possums",
   xlab="Season",
   space=c(0.15, 1),
   ylim=c(0, 80),
   las=1)
grid(ny=NA, nx=8)

out <- barplot( t(Counts), 
   beside=TRUE,
   ylab="Number of possums",
   xlab="Season",
   main="Bar chart of possum road deaths\nby season and sex of possum",
   space=c(0.15, 1),
   ylim=c(0, 80),
   las=1,
   add=TRUE)

mtext( "?", at=out[1,], side=1, line=0)
mtext( "M", at=out[2,], side=1, line=0)
mtext( "F", at=out[3,], side=1, line=0)



barplot( Counts, 
   beside=TRUE,
   xlab="Number of possums",
   #ylab="Season",
   space=c(0.15,1),
   ylim=c(0,16),
   xlim=c(0, 80),
   horiz=TRUE,
   axes=FALSE,
   names.arg=c("","",""),
   las=1)
grid(nx=NA, ny=8)

out <- barplot( Counts, 
   beside=TRUE,
   xlab="Number of possums",
   #ylab="Season",
   main="Bar chart of possum road deaths\nby season and sex of possum",
   space=c(0.15,1),
   ylim=c(0,16),
   xlim=c(0, 80),
   las=1,
   horiz=TRUE,
   names.arg=c("","",""),
   add=TRUE)

mtext( "Au", at=out[1,], side=2, line=0, las=1)
mtext( "Wn", at=out[2,], side=2, line=0, las=1)
mtext( "Sp", at=out[3,], side=2, line=0, las=1)
mtext( "Su", at=out[4,], side=2, line=0, las=1)

mtext( "Females", at=14, side=2, line=2, las=3)
mtext( "Males",   at=8, side=2, line=2, las=3)
mtext( "Unknown", at=3, side=2, line=2, las=3)




Percentages <- t(apply(Counts,1,function(x){ round(x/sum(x)*100,1)}))

barplot( Percentages, 
   beside=TRUE,
   ylab="Percentages of possums",
   #xlab="Season",
   space=c(0.15,1),
   ylim=c(0,80),
   las=1)
grid(nx=NA, ny=8)

out <- barplot( Percentages, 
   beside=TRUE,
   ylab="Percentages of possums",
   #xlab="Season",
   main="Bar chart of possum road deaths\nby season and sex of possum",
   space=c(0.15,1),
   ylim=c(0,80),
   las=1,
   add=TRUE)

mtext( "Au", at=out[1,], side=1, line=0)
mtext( "Wn", at=out[2,], side=1, line=0)
mtext( "Sp", at=out[3,], side=1, line=0)
mtext( "Su", at=out[4,], side=1, line=0)









barplot( Percentages, 
   beside=TRUE,
   ylab="Percentages of possums",
   #xlab="Season",
   space=c(0.15,1),
   ylim=c(0,80),
   las=1)
grid(nx=NA, ny=8)

out <- barplot( Percentages, 
   beside=TRUE,
   ylab="Percentages of possums",
   #xlab="Season",
   main="Bar chart of possum road deaths\nby season and sex of possum",
   space=c(0.15,1),
   #horiz=TRUE,
   ylim=c(0,80),
   las=1,
   add=TRUE)

mtext( "Au", at=out[1,], side=1, line=0)
mtext( "Wn", at=out[2,], side=1, line=0)
mtext( "Sp", at=out[3,], side=1, line=0)
mtext( "Su", at=out[4,], side=1, line=0)
```

Instead of a bar chart,
a dot chart (Fig. \@ref(fig:PossumDotCharts))
could also have been used.

```{r PossumDotCharts, echo=FALSE, fig.height=5, fig.width=5, fig.cap="A dot chart of the number of possums found as road kill, by sex and season"}
colnames(Counts) <- c("Unknown", "M", "F")

dotchart(Counts,
   main="Dot chart of possum road deaths\nby season and sex of possum",
	xlim=c(0, 80),
	xlab="Number of possums",
	pch=c(1, 2, 3, 4))
```

Another display is to construct a two-way table,
of counts (Table \@ref(tab:PossumRoadkillNumber))
or percentages (Table \@ref(tab:PossumRoadkillPercentages)).
Some of these are shown below.
Which one do you think best communicates the message in the data?


```{r PossumRoadkillNumber, echo=FALSE}

knitr::kable(Counts,
             align=c("r", "r", "r"),
             caption="The number of possums found as road kill, by sex and season")
```


```{r PossumRoadkillPercentages, echo=FALSE}
colnames(Percentages) <- c("Unknown", "M", "F")

knitr::kable(Percentages,
             align=c("r", "r", "r"),
             caption="The percentages of possums found as road kill by sex, for each season")
```












## Comparing 2-D and 3-D graphs

Using a third dimension when it is not needed is a bad idea:
the resulting graphs can be hard to read and misleading.
Remember: The purpose of a graph is to display the information in the
*simplest* and *clearest* possible way.
 
```{exercise}
Consider the relationship
(based on http://ravenanalytics.com/Articles/Misuse of statistics.pdf)
between:

* A treatment given  (A, B, C, D), and
* The gender of the patient (male or female).

Using Fig. \@ref(fig:Bar3D),
are more males or females cured for each treatment?
```


```{r Bar3D, echo=FALSE, fig.cap="A three-dimensional bar chart of treatment data"}
# Define a function
barplot3d <- function(heights, rows, transp="f0", theta=25, phi=25, bar.size=3, bar.space=0.5,
    col.lab=NULL, row.lab=NULL, z.lab=NULL, col.bar=c("#44ff58","#5844ff","#ff5844"), grid="white", ...) {
    # Set parameters
    cols    = length(heights)/rows
    calkdl  = (bar.size + bar.space)
    slupki  = matrix(heights, nrow=cols, ncol=rows)
    zakres  = pretty(0:ceiling(max(heights, na.rm=T)*1.1))
    odstep  = bar.space/2 + bar.size/2
    colors  = paste(col.bar, transp, sep="")
    shcols  = colors
    for (i1 in 1:length(colors)) 
    	shcols[i1] = paste("#",
			as.hexmode(round(unclass(as.hexmode(substr(colors[i1],2,3)))*0.8,0)),
			as.hexmode(round(unclass(as.hexmode(substr(colors[i1],4,5)))*0.8,0)),
			as.hexmode(round(unclass(as.hexmode(substr(colors[i1],6,7)))*0.8,0)),
            substr(colors[i1],8,9), sep="")

    # Prepare the grid for bars
    y = x = 0
    for (i1 in (1:rows)-1) y = c(y, bar.space/2+i1*calkdl, bar.space/1.99+i1*calkdl,
        bar.space/2+bar.size+i1*calkdl, bar.space/1.99+bar.size+i1*calkdl,
        bar.space+bar.size+i1*calkdl)
    for (i1 in (1:cols)-1) x = c(x, bar.space/2+i1*calkdl, bar.space/1.99+i1*calkdl,
        bar.space/2+bar.size+i1*calkdl, bar.space/1.99+bar.size+i1*calkdl,
        bar.space+bar.size+i1*calkdl)

    # Prepare the z matrix of bar heights
    z = matrix(nrow=length(x), ncol=length(y))
    for (i1 in (1:cols)-1)  for (i2 in (1:rows)-1) z[c(2:5)+5*i1,c(2:5)+5*i2] = 0
    for (i1 in (1:cols)-1)  for (i2 in (1:rows)-1) z[c(3:4)+5*i1,c(3:4)+5*i2] = slupki[i1+1,i2+1]

    # Prepare colors matrix
    fill   = matrix(nrow=length(x)-1, ncol=length(y)-1)
    for (i1 in (1:rows)-1) {
        fill[,c(2:5)+5*i1] = colors[i1+1]
        for (i2 in (1:cols)-1)  fill[c(4:5)+5*i2,c(3:5)+5*i1] = shcols[i1+1]
        }

    # Prepare area for plotting
    rys = persp(x, y, matrix(nrow=length(x), ncol=length(y)), col=fill, scale=F, theta=theta,
        phi=phi, zlim = range(zakres), lphi=44, ltheta=-10, axes=F, ...)

    # Add walls
    polygon(rbind(trans3d(0,0,0,rys), trans3d(0,0,max(zakres),rys), trans3d(0,max(y),max(zakres),rys),
        trans3d(0,max(y),0,rys)), col="#aaaaaa66")
    polygon(rbind(trans3d(0,max(y),0,rys), trans3d(0,max(y),max(zakres),rys), trans3d(max(x),
        max(y),max(zakres),rys), trans3d(max(x),max(y),0,rys)), col="#aaaaaa50")
    polygon(rbind(trans3d(0,0,0,rys), trans3d(0,max(y),0,rys), trans3d(max(x),max(y),0,rys),
        trans3d(max(x),0,0,rys)), col="#aaaaba66")

    # Add grid lines & numbers
    for (i1 in zakres) {
        lines(rbind(trans3d(0,0,i1,rys), trans3d(0, max(y),i1,rys)), lwd=2, col=grid)
        lines(rbind(trans3d(0,max(y),i1,rys), trans3d(max(x), max(y),i1,rys)), lwd=2, col=grid)
        text(trans3d(-(calkdl*cols)*0.04,0,i1,rys), labels=i1, adj=1, cex=0.9)
        }
    text(trans3d(-(calkdl*cols)*0.04,0,max(zakres)*1.05,rys), labels=z.lab, adj=0.8, cex=0.9)

    # Add ticks & text
    for (i1 in (1:cols)-1) {
        lines(rbind(trans3d((odstep+calkdl*i1),0,0,rys), trans3d((odstep+calkdl*i1),
            -(calkdl*rows)*0.05,0,rys)))
        if (!is.null(col.lab)) text(trans3d((odstep+calkdl*i1),-(calkdl*rows)*0.1,0,rys),
            col.lab[i1+1], adj=1, cex=0.9)
        }
    for (i1 in (1:rows)-1) {
        lines(rbind(trans3d(max(x),(odstep+calkdl*i1),0,rys), trans3d(max(x)+(calkdl*cols)*0.03,
            (odstep+calkdl*i1),0,rys)))
        if (!is.null(col.lab)) text(trans3d(max(x)+(calkdl*cols)*0.05,(odstep+calkdl*i1),0,rys),
            row.lab[i1+1], adj=0, cex=0.9)
        }

    # Plot the bars!
    par(new=T)
    persp(x, y, z, col=fill, scale=F, theta=theta, phi=phi, zlim = range(zakres),
        lphi=44, ltheta=-10, shade=0.4, axes=F, ...)

    invisible(rys)
}


# Now the data 
SeriesA <- c(23, 29, 20, 23)
SeriesB <- c(22, 30, 18, 22)

Data <- cbind(SeriesA, SeriesB)
rownames(Data) <- c("A", "B", "C", "D")
colnames(Data) <- c("Males","Females")



# Now plots
par(mar=c(8, 2, 0, 0) + 0.1)
barplot3d( Data, rows=2, bar.space=10, col.bar=c("#5844ff","#FFFAF0"),
           z.lab="No. subjects\ncured          ", 
           row.lab=colnames(Data), col.lab=rownames(Data) )


```

The artificial third dimension makes it hard to determine
the heights of the bars.

In contrast,
using a 2-D graph,
as in Fig. \@ref(fig:Bar2D),
makes it very clear whether or not more males or females cured for each treatment.
Remember:

> The purpose of a graph is to display the information
> in the clearest, simplest possible way,
> to help us understand the data.

The 3D graph clearly fails.

```{r Bar2D, echo=FALSE, fig.cap="A two-dimensional bar chart of treatment data"}
par(mar=c(4,4,4,2) + 0.1) # The default
barplot( t(Data), beside=TRUE, legend=FALSE, las=1, args.legend=list(x="bottom",bg="white", cex=0.75), 
        ylab="No. subjects cured",
        xlab="Treatment",
         col=c("#5844ff","#FFFAF0"))
         
par(xpd=TRUE) # Allow legend to be outside plot region
legend(4, 35, # Negative to go below axes 
	#col=c("#5844ff","#FFFAF0"),
	legend=c("Males","Females"), 
	#cex=1, 
	bg="white", 
	#pch=15, 
	fill=c("#5844ff","#FFFAF0"), 
	horiz=TRUE, bty="n")
```














## Other types of graphs


We have explored many types of graphs,
for different types of data.
But sometimes, 
other plots are useful.
Usually the best plot is one of those just described,
but sometimes the best plot is something different, even unique.
Always remember the driving principle:

> The purpose of a graph is to display the information
> in the clearest, simplest possible way,
> to help us understand the data.

Importantly, 
we need a graph that helps us answer the research question.
We look at three other types of graphs briefly.



### Geographic plots

When data are summarised over a geographic area,
plotting accordingly can be useful.

```{example}
A study examining lower-limb amputation incidence in Australia
(based on @data:Dillon2017:Geographic)
produced the graphs in
\@ref(fig:AmputationAustraliaF) and \@ref(fig:AmputationAustraliaM).
Clearly,
the incidence of amputation is higher in the NT than other parts of Australia for both females and males;
furthermore, 
males have higher incidence of amputation than females in every state/territory.
```

```{r AmputationAustraliaF, echo=FALSE, fig.cap="Age-adjusted incidence of lower limb amputations in Australia, from August 2007 to December 2011: females"}
library(ggplot2)
load("Data/ozdata.rda")

ozdata$Females <- ifelse(ozdata$state=="NT", "55 to 75", "15 to 35")
ggplot() + geom_polygon(data=ozdata, aes(x=long, y=lat, group=group, fill=Females) ) + 
  coord_equal() +
  scale_fill_manual(values=c("wheat", "orange"))
```

```{r AmputationAustraliaM, echo=FALSE, fig.cap="Age-adjusted incidence of lower limb amputations in Australia, from August 2007 to December 2011: males"}

ozdata$Males <- ifelse(ozdata$state=="NT", "95 to 115", "35 to 55")

ggplot() + geom_polygon(data=ozdata, aes(x=long, y=lat, group=group, fill=Males) ) + 
  coord_equal() +
  scale_fill_manual(values=c("gold", "red"))


```




### Time plots
   
Sometimes,
one of the variables is measured over *time*.
Then we can use a *time plot*,
which show how a  variable changes over time.

```{example}
In a study of green building practices~\cite{data:Paul2008:Comfort},
researchers studied the temperature in office buildings,
and produced this line graph:
	

We can see the day-to-day fluctuations,
but also 
```


```{exercise}
A study of motorcyclists fatalities~\cite{data:Haque209:MotorcycleCrashes}
included this line plot:
	

What do we learn?

```









### Case-profile plots

Sometimes the same variable is measured on each
units of analysis more than once
(i.e.\ many observations per unit of analysis).
Then we can use *case profile plots*:
plots that show how the response variable
changes for each unit of analysis
(but these can be tricky in SPSS).
Alternatively, 
and easier in SPSS,
is to plot a histogram of the *differences*
or the *changes*.
Examples of this type of data include:

* Before and after measurements.
* Measurements taken from a person's left eye and their right eye.

In both cases,
the data are called *paired*
as each unit of analysis gets a pair of similar measurements.





```{example}
A study of children with atopic asthma
[@data:Lothian2006:Whey]
included 
the graph in
Fig. \@ref(fig:PairedGraphCaseProfile).
Each line on the graph represents one person.
This type of graph is called a *case-profile plot*.


```{r PairedGraphCaseProfile, echo=FALSE, fig.cap="A case-profile plot"}
Pre <-  c( 83, 292, 293, 623, 792, 1543, 1668, 1960, 2877, 2961, 5504)
Post <- c( 83, 292, 292, 542, 709, 1000, 1000, 1626, 2502, 2711, 4504)


WhichTime <- c(
  rep(1, length(Pre)),
  rep(2, length(Post))
)

plot( c(Pre, Post) ~ WhichTime,
      pch=19,
      las=1,
      xlab="Time",
      ylab="IgE (in micrograms/L)")

for (i in 1:length(Pre)){
  lines( c(1, 2),
         c(Pre[i], Post[i]),
         col=grey(0.6)
  )
}
```

An alternative way to present the 
data---that is often better---is to 
produce a histogram of the *changes* for each person.
(This type of graph is usually easier to produce than a case-profile plot in SPSS too.)

Consider the person in the case-profile plot
whose line is at the top of the plot.
Their 'pre' IgE level is about 5500 micrograms/L,
and their 'post' IgE level is about 4500 micrograms/L,
which is a *change* (or more usefully, a *reduction*) of about 1000 micrograms/L.
We could compute these reductions for each person,
and draw a histogram of these reduction
(with one reduction for each person),
as shown in 
Fig. \@ref(fig:PairedGraphHistDiffs).


```{r PairedGraphHistDiffs, echo=FALSE, fig.cap="An alternative to a case-profile plot: A histogram of the differences"}
Diff <- Post - Pre
Reduction <- Pre - Post

hist(Reduction,
     col="wheat",
     las=1,
     breaks = seq(0, 1000, by=100),
     ylab="Number of subjects",
     xlab="Reduction in IgE (in micrograms/L)",
     main="Reduction in IgE\n pre and post supplementation")
box()
```





```{example}
A study [@data:Dawson2017:BirthdayCakes]
examined the number of bacteria on birthday cakes
*before* and *after* blowing out the candles.
Each cake has two measurements; before and after blowing of candles.
We could compute the *change* in the number of bacteria for each cake,
and plot a histogram of the difference.

CHECK! don't think this is paired
```




```{example}
A study of paced breathing~\cite{data:Pinna2005:PacedBreathing} 
included this case-profile plot:
	

The main message here is about the variation:
   Breathing frequency is more consistent (less variation)
   with paced breathing compared to spontaneous breathing.
```





## Summary, and final notes on constructing graphs
   
Always remember:

> The purpose: to display the information
> in the clearest, simplest possible way,
> to help us understand the data.

You should be able to construct the graphics by hand,
but we will generally use SPSS.
Importantly,
given the purpose of a graph,
we should understand what it communicates.
Graphs need to be clear and well-labelled with captions.

Ensure you *do not*:

*	Add artificial third dimensions.
*	Add optical optical illusions.
*	Make any errors.

Ensure that you *do*:

* Add units of measurement.
* Add titles and axis labels.
* Ensure the axis scales are appropriate.
* Add any necessary explanations.

	
	
	

To summarise one variable, 
we generally can usually use these graphs:   

* For *one quantitative variable*:
    - Histogram
    - Stem-and-leaf
    - Dot

* For *one qualitative variable*:
    - Bar chart
    - Pie chart
    - Dot

For two variables,
we can usually use these graphs
(Table \@ref(tab:GraphsToUse)).

Table: (\#tab:GraphsToUse)  The graph to use for most situations

~                         | Explanatory: Qualitative | Explanatory: Quantitative
--------------------------+--------------------------+-----------------------
Response: Qualitative     | Stacked bar chart; side-by-side bar chart             | Boxplot (horizontal); 2-D dot chart
Response: Qualitative     | Dot chart; boxplot (vertical); 2-D dot chart             	 |              

In general, the 'response' variable is on the vertical axis





```{exercise}
Critique this graph

```



```{exercise}
Critique this graph
```



```{exercise}
Critique this graph
```







```{exercise}
A study~\cite{data:Danielsson2014:Depression} examined the change in MADRS 
(a *quantitative* scale measuring level of depression)
and treatment group
(whether each person was treated using:
exercise; body awareness; or advice).
What graphs would be useful in exploring these data?

* Response variable:
* Explanatory variable:
* Relationship between these:

```











## The NHANES data: Graphical summaries

To see how these graphs can help,
we will work with the following RQ:

> Among Americans,
> is there a relationship between smoking status, 
> and direct HDL cholesterol levels?

From the RQ,
we see that:

* Population:  Americans.
* Outcome: Direct HDL cholesterol levels.
* Comparison/connection: Between current smoking statuses: Current smoker or not
* Intervention: (No intervention).

This is a *relational RQ*,
and can be answered using an *observational study*.


```{exercise}
As with any study,
we should think about managing confounding by
thinking about possible extraneous variables that we could measure or observe;
can you think of any?   
```

In this study,
clearly we cannot collect primary data.
However,
data to answer this (and many other) RQs
is obtained from the 
American *National Health and Nutrition Examination Survey* (NHANES)~\cite{data:NHANES3:Data,data:NHANES3,data:NHANES:Rpackage}.
From the 
[NHANES webpage](http://www.cdc.gov/nchs/nhanes/about_nhanes.htm), 
this survey:

> ... examines a nationally *representative* sample of about 5,000 persons each year\dots 
> located in counties across the country, 15 of which are visited each year.
>
> --- [NHANES webpage](http://www.cdc.gov/nchs/nhanes/about_nhanes.htm)

The data we have available is equivalent to a 
"a simple random sample from the American population"~\cite{data:NHANES:Rpackage}.
We have 10\,000 observations on scores of variables
(from the 2009/2010 and the 2011/2012 surveys).
      
```{exercise}
From the description,
what sampling scheme seems likely in the NHANES study?
```   





```{exercise}
Describe the variables from the NHANES study in
Table \@ref(tab:ClassifyNHANES).

Table: (\#tab:ClassifyNHANES)  Classifying the NHANES variables

Variable              | Quantitative: (discrete/continuous?) or qualitative (nominal/ordinal)?  | In SPSS? 
----------------------+-------------------------------------------------------------------------+----------
HDL cholesterol       | | 
Current smoker?       | |
Age                   | |
Sex                   | |  
Body weight           | |

```   
   

Any time we have a RQ,
it is sensible to explore and understand the data,
and using graphs is a great way to do so.

```{exercise}
We need to understand the response variable.
Here are some of the observations:
How would you summarise these...
   
PLOT

It is hard...  but graphs can help.
```


We usually begin by graphing both the response and explanatory variables
involved in the RQ.


For the NHANES data,
the response variable is direct HDL cholesterol,
which is quantitative continuous;
the histogram is shown in
Fig.~\ref{FG:GRAPHS:NHANEShistogram}.
Just over 500 people had a HDL cholesterol level 200 or greater but under 225;
 just over 100 had a level 150 or greater but under 175.


```{r echo=FALSE}
hist(NHANES$DirectChol,
	xlab="Direct HDL cholesterol (mmol/L)",
	ylab="Frequency",
	las=1,
	ylim=c(0, 4500),
	xlim=c(0, 4.5),
	main="",
	breaks=seq(0, 4.5, by=0.5),
	col="wheat")
box()
```

   
The explanatory variable is 'current smoking status',
which we could graph using a bar chart:
               
```{r NHANES-Boxplot-Smoke, echo=FALSE}
barplot( table(NHANES$SmokeNow),
	xlab="Does respondent currently smoke?",
	ylab="Count",
	ylim=c(0, 2000),
	las=1,
	col="wheat")
box()
```

But this only conveys *one* piece of information,
so a graph probably isn't really  necessary:
That `r round( table(NHANES$SmokeNow)["Yes"] / sum( table(NHANES$SmokeNow) ) * 100, 1)`\% 
(just under half)
of the respondents were current smokers.
Recall that there were 10\,000 respondents,
yet the bar chart clearly does not have 10\,000 responses.
Many respondents did not answer this question.


The main RQ here involves the relationship between 
direct HDL cholesterol and current smoking statu,s
so we need a graph that displays this relationship of interest:
a boxplot.
      
```{r NHANES-boxplot, echo=FALSE}
boxplot(NHANES$DirectChol ~ NHANES$SmokeNow, 
	col="wheat", 
	las=1,
	ylim=c(0, 4),
	ylab="Direct HDL cholesterol (mmol/L)",
	xlab="Does respondent smoke now?"
)
```

From the plot,
does there appear to be much difference in the average HDL cholesterol
concentrations between the two smoking groups?


It is always useful to explore relationships between the
response variable and extraneous variables,
and between the
explanatory variable and extraneous variables.
This is especially the case with observational studies like the NHANES study,
because confounding can be a problem.
We will plot the response and explanatory variables 
against weight and gender
(Fig.~\ref{FG:GRAPHS:NHANESConfounders})
to explore potential confounding.



```{r NHANES-plots, echo=FALSE}
par( mfrow=c(2, 2))

plot( DirectChol ~ Weight, data=NHANES, 
	xlab="Weight (in kg)", 
	ylab="Direct HDL cholesterol (mmol/L)",
	ylim=c(0, 4),
	las=1)

plot( Weight ~ SmokeNow, data=NHANES, 
	xlab="Current smoker?", 
	ylab="Current smoking status",
	horiz=TRUE,
	ylim=c(0, 250),
	las=1)

plot( DirectChol ~ Gender, data=NHANES, 
	xlab="Gender", 
	ylab="Direct HDL cholesterol (mmol/L)",
	names=c("Female", "Male"),
	ylim=c(0, 4),
	las=1)

par(mar=c(5, 4, 4, 6) + 0.1)  # DEFAULT: c(5, 4, 4, 2) + 0.1.
barplot( xtabs(~ SmokeNow + Gender, data=NHANES), 
	ylim=c(0, 2000),
	xlab="Gender", 
	ylab="Number of people",
	names=c("Female", "Male"),
	col=c("red", "lightblue"),
	las=1)
text(2,  500, "Smokers"  	 )
text(2, 1500, "Non-smokers")
```




```{exercise}
What do we learn from these graphs?
```




