# (PART) Reporting, writing and reading research {-}








# Reading research


```{r echo=FALSE}
library(knitr)
# set global chunk options
options(formatR.arrow=TRUE,width=90)
```


```{r echo=FALSE}
SixSteps(6, "Reading")
```

## Introduction

AGAIN: LOOKS OLD!

Science requires us to read the research of others.
         Throughout your study at university,
         you will need to read and understand the research of others.
         Research is communicated in *journal articles*
         and *presentations*
         (conferences; seminars).
         Thousands of journal articles are in the library
         (many online).
         We need to understand the language of research 
         to understand these articles.

What do we look for when we read a paper?

* What is the *research question*? (Or: aims, objectives, purpose, hypotheses, \dots)
* Is the study *well-designed*? Are definitions clear?  (Consider: Hawthorne effect, blinding, \dots)
* Is the *data collection* protocol appropriate and well documented?
* Is the data *summary* appropriate,   
         complete and clear?
* Is the *analysis* appropriate and clear?
* Are the results accurate, appropriate and well *reported*?

This is the process of research we have been studying.
   


All this information is important
for understanding the research,
but the information is not always easy to find!
Reading papers can be hard work\dots
      
The Abstract 
(or Summary, or Overview, or \dots)
*should* be a useful overview of the whole paper (without details).

Articles should be critically evaluated:
they may not all be good just because they are published.
Ask questions about the quality of the research,
how can results be generalised,
what limitations exist, etc.      




## Example 1

IMAGE

* RQ: Can you identify P, O, C and I?
* What is the design for collecting the data?
* What do we learn about the data collection?
* How does the sample data answer the question?
* Is this a one- or two-tailed RQ?
* What are the limitations of this study?
* To what population do the results generalise?










## Example 2


IMAGE      




* What are the variables?
* What *type* of variables are these?



## Example 3

Watch this space


# Writing research

## Introduction
```{r echo=FALSE}
SixSteps(6)
```


All of you will need to read the research of others.
Some of you will need to write about your own studies
and experiments.
You need to understand the language of research 
to be able to write scientifically.
         


Some information here is based on 
[Dr. Michael Lufaso's notes](http://www.unf.edu/~michael.lufaso/chem4931/lecture3.pdf)
and
[Prof. Tony Roberts](http://www.maths.adelaide.edu.au/anthony.roberts/LaTeX/ltxwrite.php).

A scientific paper **should**:

* Present the facts in an unbiased manner
* Be clear, concise and complete
* Use facts to make statements
* Be complete enough that other professionals can repeat your work

A scientific paper **should not**:

* Be haphazard, jumbled and illogical 
* Be used as your own personal soapbox 
* Reach conclusions not based on the reported evidence  
* Be for insiders only





[USQ Plagiarism website](http://www.usc.edu.au/University/Library/Tutorials/OnlineUndergrad/PlagiarismReferencing/Plagiarism.htm)

Research involves using other people's ideas and research to develop your own conclusions.
Acknowledge all sources you have used when writing your research, otherwise you are committing plagiarism.
   Plagiarism:
```{definition, name="Plagiarism"}
'Something appropriated and passed off as one's own'. Delbridge, A. et al, 1997, 
*The Macquarie Dictionary*, 3rd edn
```

Plagiarism is a serious offence: theft of intellectual property.
**Do not plagiarise**:

* Do not take parts of sentences or complete sentences directly from papers
* Use quotes if necessary and cite work (sparingly)
* Have someone else critically read the paper

 Plagiarism can destroy people's careers and affect the reputation and status of the University
 If you are caught plagiarising, you will be penalised.
   Penalties can range from having to resubmit assignments and being marked down,
   to failing that assignment or being failed from the course.
   In very serious cases you could be expelled from USC.





## Structure

Many scientific papers have these (or similar) sections
(though it varies a lot by discipline):

* Title and authors
* **A**bstract
* **I**ntroduction: Why did you do this?
* **M**aterials and method: What did you do?
* **R**esults: What did you find?
* **D**iscussion (or: Summary, or Conclusions): What does it mean?
* References (or: Bibliography)

Often the acronym AIMRaD or IMRaD is used to help recall these sections.

These components 
often capture certain part sof the six-step research process we have been discussing:


IMAGE!




## Writing Scientifically: Title

A title should be a clear description of what you are talking about,
and use the *fewest* words possible that cover the purpose of the paper.
(Almost) never use abbreviations in a title.
It is useful if the title can explains why your study is different from the many others that exist.
Titles sometimes
include the technique or method used.
   
   
Here is a good exmaple:

> Beauty sleep: 
> experimental study on the perceived health and attractiveness of sleep deprived people
>
> --- @data:axelsson:beautysleep



Here is a *poor* example:

> The nucleotide sequence of a 3.2 kb segment of mitochondrial maxicircle DNA from Crithidia fasciculata containing the gene for 
> cytochrome oxidase subunit III, the N-terminal part of the apocytochrome b gene and a possible frameshift gene; further evidence 
> for the use of unusual initiator triplets in trypanosome mitochondria
> 
> --- @sloof1987nucleotide

 
This example is *even worse*:

> Reaction of a bidentate ligands (4,4'-dimethyl 2,2'-bipyridine) with planar-chiral chloro-bridged ruthenium: 
> Synthesis of cis-dicarbonyl[4,4'-dimethyl-2,2'-bipyridine- $\kappa$ O1, $\kappa$ O2]
> $\lbrace$ 2-[tricarbonyl($\eta$ 6-phenylene- $\kappa$ C1)chromium]pyridine- $\kappa$N $\rbrace$ ruthenium hexafluorophosphate
>
> --- @hijazi2013

This example is even worse,
since the article needed to be retracted due to unethical conduct of one author
[@hijazi2013retracted].


## Writing Scientifically: Abstract

A short paragraph which summarises the *whole* paper;
it is not an introduction!
A good abstract:

* Summarises the whole paper;
* Gives a concise statement that describes the purpose of the paper
   (so give the RQ and/or hypotheses);
* Includes results and conclusion (specific but not detailed);
*  Does not include anything that is not in the paper.

Writing  the abstract after the paper is fully written is usually sensible.


## Writing Scientifically: Introduction

The introduction has two purposes:

1. To set up the context for the paper;
2. Allow the reader to become familiar with the theoretical groundwork of the subject.

Provides a clear statement of the study's RQ
(sometimes called the purpose, aim, objective, hypotheses, etc.).
It introduces the topic,
states the relevance of the topic,
defines and explains the terms, acronyms, concepts, and theories necessary to understand the paper.
It is as short and complete as possible.


The introduction usually includes a literature review too,
but we do not require a literature review in this course.




## Writing Scientifically: Materials and methods


The Materials and Methods section:

* Describes how the *sample* was obtained;
*  Describes {how} the data were *collected*: 
   the data collection *protocol*;
* Describes how the data were *analysed*
   (including the software used, and the statistical methods used);
* Describe any specialized equipment used
   (don't list pencils, rulers, paper, etc.!).






## Writing Scientifically: Results

The Results section:

* Shows all the relevant findings from the research;
* Presents a summary of the dataset: 
      the number of cases, the number of missing values, 
      and a verbal description of all variables;
* Presents tabular, numerical or graphical summary of each variable;
* Gives a brief verbal interpretation of these summaries;
* Gives the *results* from any hypothesis tests and confidence intervals;
* Identifies trends, consistencies, anomalies etc.;
* Does *not* contain an interpretation or explanation of the results of the tests; 
      that is the purpose of the Discussion;
* Defines any notation you use.







## Writing Scientifically: Discussion

The Discussion section:

* Summarises the results;
* Gives a short evaluation of the results:
   Be concise;
* Refers back to the stated RQ;
* Discusses limitations,  problems, challenges;
* Tries to anticipate and respond to potential questions about the research.


Don't try to sell your conclusions (the reader should reach their conclusions based on the facts you presented).




## Figures and tables

All tables and figures need a caption:

* Tables typically have captions *above*
* Figures typically have captions *below*

Captions should be, as far as possible,
self-contained
   
Tables have very few lines:
Vertical lines are almost never needed
 

```{example}
An article
[@data:Baur2012:Firefighters]
includes this in the Abstract:

> Cardiovascular disease (CVD) accounts for 45\% of on-duty fatalities among firefighters, 
> occurring primarily in firefighters with excess CVD risk factors in patterns resembling the metabolic syndrome (MetSyn).    
> Additionally, firefighters have a high prevalence of obesity and sedentary behavior suggesting that MetSyn is also common.
> Therefore we assessed the prevalence of MetSyn in firefighters 
> and its association with cardiorespiratory fitness (CRF) 
> in a cross-sectional study of 957 male career firefighters.
>
> --- @data:Baur2012:Firefighters, p. 2331

The paper includes this table; critique this table.

IMAGE

The paper includes this figure; critique this figure.

IMAGE

```

```{example}
Another study
[@data:Baughman2007:SixMinuteWalk]
gave this information:

> The aim of our study was to determine the range of 6MWD [6-minute walk distance]
> in an unselected group of sarcoidosis patients.
> We performed a prospective study of sarcoidosis patients followed up in one tertiary sarcoidosis clinic.
>
> --- @data:Baughman2007:SixMinuteWalk, p. 208

Critique this graph that appears in the paper:
	
IMAGE

```



## General writing tips

TONY ROBERST WEB SITE




## Other things



* References:
    - Give the full citations of work you have referenced, in the correct format;
    - Organise and format your references correctly:
      Many disciplines or journals have very strict guidelines
      for how references must be listed;
    
* Acknowledgements:
   Thank people who legitimately contributed to your report,
   and research funding bodies.
* Appendices
    - In the appendices,
      place important material that
      would break the flow of the narrative.
    - Appendices may include large tables, images, detailed discussions of technical details, \dots




## Style

TONY'S STUFF IN HERE??


Different disciplines have their own styles.
Use short sentences;
be unambiguous:

> Don't write so that you can be understood, 
> write so that you can't be misunderstood.
> 
> --- Attributed to William Howard Taft

Many disciplines prefer you to write using passive voice,
and many disciplines prefer you to wrote in the past tense.
In addition,
many disciplines prefer authors to avoid first person singular/plural  (''I did this\dots'') wherever possible. 
In all cases,
use inclusive language.

And finally:

* Use, but do not rely upon, the spell checker;
* Don't talk down to the reader;
* Omit any words, phrases, sentences that add nothing to the paper;
*   Ensure correct content, grammar, spelling, format;
* Check for commonly misused words: there/their; your/you're;
* Ensure capitalisation is correct;
* Sentences should say what they mean,
   and mean what they say;
* Use apostrophes correctly.












## Example

Consider the NHANES data again.
In preparing a paper about this study,
suppose the following graph and tables were produced.
Critique these.





```{r echo=FALSE}
library(NHANES)
data(NHANES)

plot(BPSys1 ~ SmokeNow, data=NHANES, 
	xlab="",
	col="purple") 
```


Group                |  Mean     | Std dev
---------------------+-----------+---------------
Current smoker       |  206.6    | 46    
Current non-smoker   |  214.64   | 48.79  
*Difference*         |    8.03   |
95\% CI              | 1.25      | 14.8 




